### 2024-09-23 paper/paper.qmd
paper setup

---
title: "My title"
subtitle: "My subtitle"
author: 
  - Ziheng Zhong
thanks: "Code and data are available at: https://github.com/iJustinn/Toronto_Cycling_Network"
date: today
date-format: long
abstract: "My abstract"
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

```{r}
#| label: fig-one
#| fig-cap: one
#| echo: false

```

```{r}
#| label: fig-two
#| fig-cap: two
#| echo: false
#| warning: false
#| message: false

```

# Results

```{r}
#| echo: false
#| eval: true
#| label: tbl-one
#| tbl-cap: "cap"
#| warning: false

```

Our results are summarized in @tbl-modelresults.

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

\newpage

# References

### 2024-09-23 paper/paper.qmd
paper sourse and measurement added

@@
 format: pdf
+toc: true
+toc-depth: 3
 number-sections: true
@@
 
-library(tidyverse)
-library(palmerpenguins)
+# Load packages
+package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr")
+
+# Function to check and install missing packages
+install_and_load <- function(package_list) {
+  for (package in package_list) {
+    if (!require(package, character.only = TRUE)) {
+      install.packages(package)
+      library(package, character.only = TRUE)
+    }
+  }
+}
+
+install_and_load(package_list)
+```
+
+```{r}
+#| include: false
+#| warning: false
+#| message: false
+
+# Load data
+
 ```
@@
 
-The remainder of this paper is structured as follows. @sec-data....
+The remainder of this paper is structured as follows.
 
 # Data {#sec-data}
+
+## Source
+
+The data used in this paper was collected by the OpenDataToronto Library [@citeODT].
+OpenDataToronto provides a platform for the public to access various datasets related to Toronto’s civic operations and urban infrastructure.
+The specific dataset used in this research is the 'Cycling Network' [@citeDataset], which provides detailed information on Toronto’s bicycle infrastructure, including dedicated lanes, multi-use trails, and shared roadways.
+This dataset plays a crucial role in promoting active transportation and urban sustainability efforts within the city.
+It is frequently updated to reflect ongoing expansions and modifications to the cycling infrastructure, aligning with Toronto's broader efforts to reduce congestion, enhance mobility, and support environmental goals.
+This dataset not only informs local decisions but also contributes to broader discussions on urban transportation planning and sustainability in growing cities around the world.
+
+Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR].
+Also with support of additional packages in R: `sf` [], `readr` [@citeReadr], `ggplot2` [@citeGgplot], `osmdata` [], `tidyverse` [@citeTidyverse], `jsonlite` [], `dplyr` [@citeDplyr], .
+
+## Measurement
+
+The 'Cycling Network' dataset from Open Data Toronto tracks various aspects of Toronto's cycling infrastructure, such as installation dates, upgrades, street names, and the type of infrastructure (e.g., sharrows, multi-use trails).
+Each entry represents a segment of the cycling network, where real-world phenomena, such as the construction or upgrading of cycling paths, are documented.
+For example, when a cycling path is installed or upgraded, the responsible authorities collect information such as the installation year (e.g., 2001) and any subsequent upgrades (e.g., 2021).
+This data is then digitized into structured entries in the dataset.
+
+To go from a physical event (e.g., the installation of a bike lane) to a dataset entry, detailed records are maintained by the city's transportation department.
+These records are geospatially coded (as seen in the `geometry` column), allowing each cycling path to be mapped precisely in relation to other city infrastructure.
+Thus, the phenomena of constructing or updating a bike lane becomes an entry with attributes such as the street name, road class, and installation history.
+This allows researchers to analyze trends in cycling infrastructure and its expansion over time.
+
+## Method {#sec-data-method}
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-one
+#| tbl-cap: "cap"
+#| warning: false
+
+```
+
+# Results {#sec-res}
+
+## Data Trend
 
@@
 
-# Results
+## Distributions
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-one
-#| tbl-cap: "cap"
-#| warning: false
-
-```
+## Maps
 
@@
 
-# Discussion
+# Discussion {#sec-dis}
 
-## First discussion point {#sec-first-point}
+## First discussion point
 
@@
 
-# Appendix {-}
+# Appendix {#sec-app}
 

### 2024-09-23 scripts/02-data_cleaning.R
data cleaning updated

@@ -126,7 +126,7 @@ write_csv(
 
 
 
-#### Clean data 3 ####
+#### Clean data 4 ####
 # Lane Type data
 # Load the dataset
 data <- read_csv("data/raw_data/raw_data.csv")

### 2024-09-23 paper/paper.qmd
paper data section updated

@@
 # Load packages
-package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr")
+package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr", "here", "knitr", "kableExtra")
 
@@
 install_and_load(package_list)
-```
-
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-# Load data
-
 ```
@@
 
-Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR].
-Also with support of additional packages in R: `sf` [], `readr` [@citeReadr], `ggplot2` [@citeGgplot], `osmdata` [], `tidyverse` [@citeTidyverse], `jsonlite` [], `dplyr` [@citeDplyr], .
+Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages which will be talked about in the @sec-data-method.
 
@@
 To go from a physical event (e.g., the installation of a bike lane) to a dataset entry, detailed records are maintained by the city's transportation department.
-These records are geospatially coded (as seen in the `geometry` column), allowing each cycling path to be mapped precisely in relation to other city infrastructure.
+These records are geospatially coded (column geometry in this dataset), allowing each cycling path to be mapped precisely in relation to other city infrastructure.
 Thus, the phenomena of constructing or updating a bike lane becomes an entry with attributes such as the street name, road class, and installation history.
@@
 
+Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR].
+Also with support of additional packages in R: `sf` [], `readr` [@citeReadr], `ggplot2` [@citeGgplot], `osmdata` [], `tidyverse` [@citeTidyverse], `jsonlite` [], `dplyr` [@citeDplyr], `here` [], `knitr` [], `kableExtra` [], `` [], `` [], .
+
 ```{r}
@@
 #| eval: true
-#| label: tbl-one
-#| tbl-cap: "cap"
+#| label: tbl-raw-coordinates
+#| tbl-cap: "Raw Coordinates"
 #| warning: false
 
+# Load the dataset
+data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+
+# Extract the first part of the geometry string for each row
+data_geometry <- data %>%
+  mutate(simplified_geometry = substr(geometry, 1, 92)) %>%  # Extract the first 100 characters
+  select(simplified_geometry)
+
+# Display the simplified geometry data
+head(data_geometry) %>%
+  kable("latex", booktabs = TRUE, col.names = c("Simplified Geometry")) %>%
+  kable_styling(latex_options = c("hold_position")) %>%
+  column_spec(1, width = "40em")  # Adjust the width as needed
 ```
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-clean-coordinates
+#| tbl-cap: "Extracted and Cleaned Coordinates"
+#| warning: false
+
+# Load the dataset
+data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+
+# Function to extract and clean coordinates from geometry
+extract_coordinates <- function(geo) {
+  if (!is.na(geo)) {
+    geo_clean <- gsub("'", "\"", geo)
+    json_data <- tryCatch({
+      fromJSON(geo_clean)
+    }, error = function(e) {
+      return(NULL)
+    })
+
+    if (!is.null(json_data) && !is.null(json_data$coordinates)) {
+      coords <- unlist(json_data$coordinates, recursive = TRUE)
+      longitude <- coords[grepl("^-79", coords)]
+      latitude <- coords[grepl("^43", coords)]
+      if (length(longitude) == length(latitude)) {
+        coords_matrix <- cbind(longitude, latitude)
+        return(as.data.frame(coords_matrix))
+      }
+    }
+  }
+  return(NULL)
+}
+
+# Extract coordinates and clean data
+all_coordinates <- lapply(data$geometry, extract_coordinates)
+valid_coordinates <- all_coordinates[!sapply(all_coordinates, is.null)]
+data$id <- seq_len(nrow(data))
+coordinates_data <- do.call(rbind, Map(function(coords, id) {
+  coords$id <- id
+  return(coords)
+}, valid_coordinates, data$id[!sapply(all_coordinates, is.null)]))
+
+# Set column names and remove NA values
+colnames(coordinates_data) <- c("Longitude", "Latitude", "ID")
+coordinates_data <- coordinates_data %>% filter(!is.na(Longitude), !is.na(Latitude))
+
+# Display the first few rows of the cleaned coordinate data
+head(coordinates_data) %>%
+  kable("latex", booktabs = TRUE, col.names = c("Longitude", "Latitude", "ID"))
+```
+
+In the first data cleaning process, the focus was on extracting and cleaning geographical coordinates from the 'geometry' column, which contains location data in JSON-like format, @tbl-raw-coordinates displays how the data was like.
+A custom function was created to parse the coordinates, filter out invalid entries, and extract the longitude and latitude values.
+The extracted coordinates were combined into a new dataset shon in @tbl-clean-coordinates, with unique IDs to match the original entries.
+Any rows with missing coordinate data were removed, and the cleaned data was saved into a separate file for further analysis.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-clean-upgraded
+#| tbl-cap: "Number of Lane Upgrades by Year"
+#| warning: false
+
+# Load the dataset
+data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+
+# Step 1: Select the 'UPGRADED' column
+upgraded_data <- data %>%
+  select(UPGRADED)
+
+# Step 2: Filter out invalid and missing years
+upgraded_data <- upgraded_data %>%
+  filter(!is.na(UPGRADED) & UPGRADED > 1800)  # Keep valid years only
+
+# Step 3: Convert the 'UPGRADED' column to numeric
+upgraded_data <- upgraded_data %>%
+  mutate(UPGRADED = as.numeric(UPGRADED))
+
+# Step 4: Group the data by the year of upgrade and summarize the number of upgrades
+upgraded_summary <- upgraded_data %>%
+  group_by(UPGRADED) %>%
+  summarise(num_lanes = n())
+
+# Step 5: Display the summarized data in a table
+tail(upgraded_summary) %>%
+  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Upgraded Lanes"))
+```
+
+The second cleaning process dealt with the 'UPGRADED' column, which contains information about the years when cycling lanes were upgraded.
+Invalid or missing years were filtered out, and the remaining valid data was converted to a numeric format.
+The dataset was then grouped by the year of upgrade, with a summary created to count the number of lanes upgraded in each year.
+This summarized data, @tbl-clean-upgraded, was saved for use in analyzing trends in lane upgrades over time.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-clean-installed
+#| tbl-cap: "Number of Bikeways Installed by Year"
+#| warning: false
+
+# Load the dataset
+data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+
+# Step 1: Select the 'INSTALLED' column
+installed_data <- data %>%
+  select(INSTALLED)
+
+# Step 2: Filter out invalid and missing years
+installed_data <- installed_data %>%
+  filter(!is.na(INSTALLED) & INSTALLED > 1800)  # Keep valid years only
+
+# Step 3: Convert the 'INSTALLED' column to numeric
+installed_data <- installed_data %>%
+  mutate(INSTALLED = as.numeric(INSTALLED))
+
+# Step 4: Group the data by the year of installation and summarize the number of installations
+installed_summary <- installed_data %>%
+  group_by(INSTALLED) %>%
+  summarise(num_bikeways = n())
+
+# Step 5: Display only the first 6 rows of the summarized data in a table
+head(installed_summary) %>%
+  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Installed Bikeways"))
+```
+
+In the third cleaning process, the focus was on the 'INSTALLED' column, which records the years when cycling infrastructure was first installed.
+Similar to the 'UPGRADED' column, invalid or missing years were removed, and the data was converted to numeric format.
+The installation data was grouped by year to count the number of bikeways installed each year.
+The resulting summary, @tbl-clean-installed, was saved for further trend analysis on the growth of the cycling network.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-clean-comfort
+#| tbl-cap: "Classification of Cycling Lanes by Comfort Level"
+#| warning: false
+
+# Load the dataset
+data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+
+# Step 1: Select relevant columns for classification
+lane_type_data <- data %>%
+  select(INFRA_HIGHORDER, INFRA_LOWORDER, INSTALLED, UPGRADED)
+
+# Step 2: Filter out rows with missing values in the relevant columns
+lane_type_data <- lane_type_data %>%
+  filter(!is.na(INFRA_HIGHORDER) & !is.na(INFRA_LOWORDER))
+
+# Step 3: Classify lanes into comfort levels based on infrastructure types
+lane_type_data <- lane_type_data %>%
+  mutate(
+    Comfort_Level = case_when(
+      str_detect(INFRA_HIGHORDER, "Protected|Multi-Use") | str_detect(INFRA_LOWORDER, "Protected|Multi-Use") ~ "High Comfort",
+      str_detect(INFRA_HIGHORDER, "Bike Lane") | str_detect(INFRA_LOWORDER, "Bike Lane") ~ "Moderate Comfort",
+      TRUE ~ "Low Comfort"
+    )
+  )
+
+# Step 4: Select relevant columns to display (Comfort Level, Installed, and Upgraded year)
+final_lane_data <- lane_type_data %>%
+  select(INSTALLED, UPGRADED, Comfort_Level)
+
+# Step 5: Display the first 6 rows of the cleaned and classified data
+head(final_lane_data) %>%
+  kable("latex", booktabs = TRUE, col.names = c("Installed Year", "Upgraded Year", "Comfort Level"))
+```
+
+Finally, the fourth cleaning process involved classifying cycling lanes based on the type of infrastructure.
+Data from the 'INFRA_HIGHORDER' and 'INFRA_LOWORDER' columns was cleaned, and missing values were removed.
+The lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, depending on whether the infrastructure was protected or involved bike lanes.
+The classification, along with year data for installation and upgrades, was saved for analysis on the types of cycling lanes and their comfort levels.
+@tbl-clean-comfort shows how it looks like.
 

### 2024-09-23 scripts/04-charts.R
charts updated

@@ -259,8 +259,6 @@ ggsave("other/charts/moderate_comfort_upgrade_distribution.jpg")
 # Upgraded of High Comfort
 # Load the data
 lane_type_data <- read_csv("data/analysis_data/lane_type_data.csv")
-# Load the cleaned lane type data
-lane_type_data <- read_csv("data/analysis_data/lane_type_data.csv")
 
 # Filter data for 'High Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
 high_comfort_data <- lane_type_data %>%

### 2024-09-23 paper/paper.qmd
paper charts added

@@
   column_spec(1, width = "40em")  # Adjust the width as needed
+
 ```
@@
   kable("latex", booktabs = TRUE, col.names = c("Longitude", "Latitude", "ID"))
+
 ```
@@
   kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Upgraded Lanes"))
+
 ```
@@
   kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Installed Bikeways"))
+
 ```
@@
   kable("latex", booktabs = TRUE, col.names = c("Installed Year", "Upgraded Year", "Comfort Level"))
+
 ```
@@
 ```{r}
-#| label: fig-one
-#| fig-cap: one
+#| label: fig-installed
+#| fig-cap: "Toronto Bikeways Installed by Year"
+#| fig-subcap: ["Number", "Trend"]
 #| echo: false
+#| warning: false
+#| message: false
+#| layout-ncol: 2
+#| layout-nrow: 1
+
+# Load the data
+installed_data <- read_csv(here("data", "analysis_data", "installed_data.csv"))
+
+# Generate the bar chart for "Number of Bikeways Installed by Year"
+ggplot(installed_data, aes(x = INSTALLED, y = num_bikeways)) +
+  geom_bar(stat = "identity", fill = "darkgrey") +  # Bar chart
+  labs(x = "Year of Installation",
+       y = "Number of Bikeways") +
+  scale_x_continuous(breaks = seq(min(installed_data$INSTALLED), max(installed_data$INSTALLED), by = 1)) +  # Display all years
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability
+
+# Generate the line chart with trend line for "Number of Bikeways Installed by Year"
+ggplot(installed_data, aes(x = INSTALLED, y = num_bikeways)) +
+  geom_line(aes(group = 1), color = "black", size = 1) +  # Adds a line connecting each point
+  geom_point(color = "black", size = 2) +  # Adds points at each data value
+  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Adds a linear trend line
+  labs(x = "Year of Installation",
+       y = "Number of Bikeways") +
+  scale_x_continuous(breaks = seq(min(installed_data$INSTALLED), max(installed_data$INSTALLED), by = 1)) +  # Display all years
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability
 
@@
 ```{r}
-#| label: fig-two
-#| fig-cap: two
+#| label: fig-upgraded
+#| fig-cap: "Toronto Bikeways Upgraded by Year"
+#| fig-subcap: ["Number", "Trend"]
 #| echo: false
@@
 #| message: false
+#| layout-ncol: 2
+#| layout-nrow: 1
+
+# Load the data
+upgraded_data <- read_csv(here("data", "analysis_data", "upgraded_data.csv"))
+
+# Generate the bar chart for "Number of Bikeways Upgraded by Year"
+ggplot(upgraded_data, aes(x = UPGRADED, y = num_lanes)) +
+  geom_bar(stat = "identity", fill = "darkgrey") +  # Bar chart
+  labs(x = "Year of Upgrade",
+       y = "Number of Bikeways") +
+  scale_x_continuous(breaks = seq(min(upgraded_data$UPGRADED), max(upgraded_data$UPGRADED), by = 1)) +  # Display all years
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability
+
+# Generate the line chart with trend line for "Number of Bikeways Upgraded by Year"
+ggplot(upgraded_data, aes(x = UPGRADED, y = num_lanes)) +
+  geom_line(aes(group = 1), color = "black", size = 1) +  # Adds a line connecting each point
+  geom_point(color = "black", size = 2) +  # Adds points at each data value
+  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Adds a linear trend line
+  labs(x = "Year of Upgrade",
+       y = "Number of Bikeways") +
+  scale_x_continuous(breaks = seq(min(upgraded_data$UPGRADED), max(upgraded_data$UPGRADED), by = 1)) +  # Display all years
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for readability
 
@@
 
+```{r}
+#| label: fig-comfort
+#| fig-cap: "Toronto Bikeway Comfort Level Distribution"
+#| fig-subcap: ["All levels", "Low Comfort", "Moderate Comfort", "High Comfort"]
+#| echo: false
+#| warning: false
+#| message: false
+#| layout-ncol: 2
+#| layout-nrow: 2
+
+# Load the data
+lane_type_data <- read_csv(here("data", "analysis_data", "lane_type_data.csv"))
+
+# Combine the 'INSTALLED' and 'UPGRADED' years into one column for better visualization
+lane_type_years <- lane_type_data %>%
+  pivot_longer(cols = c(INSTALLED, UPGRADED), names_to = "Year_Type", values_to = "Year") %>%
+  filter(!is.na(Year)) %>%  # Remove NA values in the 'Year' column
+  group_by(Comfort_Level) %>%
+  summarise(count = n())  # Count the occurrences of each Comfort Level
+
+# Calculate percentages
+lane_type_years <- lane_type_years %>%
+  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column
+
+# Define a more contrasting Morandi-inspired color scheme
+contrasting_morandi_colors <- c(
+  "High Comfort" = "#B39B8E",   # Muted brownish-pink
+  "Moderate Comfort" = "#8A7967", # Stronger grey-brown
+  "Low Comfort" = "#6E7062"     # Dark muted gray-green
+)
+
+# Generate the pie chart with border lines and larger text/legend
+ggplot(lane_type_years, aes(x = "", y = percentage, fill = Comfort_Level)) +
+  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
+  coord_polar("y", start = 0) +
+  geom_text(aes(label = paste0(percentage, "%")), 
+            position = position_stack(vjust = 0.5), 
+            size = 6,   # Increase text size for percentage labels
+            fontface = "bold") +  # Make percentage text bold
+  scale_fill_manual(values = contrasting_morandi_colors) +  # Apply the more contrasting Morandi colors
+  labs(fill = "Comfort Level") +
+  theme_void() +  # Remove axis, grid lines, and background
+  theme(legend.position = "right", 
+        legend.title = element_text(size = 16),  # Increase legend title size
+        legend.text = element_text(size = 14))  # Increase legend text size
+
+# Filter data for 'Low Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
+low_comfort_data <- lane_type_data %>%
+  filter(Comfort_Level == "Low Comfort") %>%
+  mutate(
+    Upgrade_Status = ifelse(is.na(UPGRADED), "Not Upgraded", "Upgraded")
+  ) %>%
+  group_by(Upgrade_Status) %>%
+  summarise(count = n())  # Count the occurrences of each category
+
+# Calculate percentages
+low_comfort_data <- low_comfort_data %>%
+  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column
+
+# Define colors for 'Upgraded' and 'Not Upgraded' categories
+upgrade_colors_low <- c(
+  "Upgraded" = "#8A7967",  # Gray-brown for upgraded
+  "Not Upgraded" = "#B39B8E"  # Muted pinkish tone for not upgraded
+)
+
+# Generate the pie chart with borders and larger text/legend for Low Comfort lanes
+ggplot(low_comfort_data, aes(x = "", y = percentage, fill = Upgrade_Status)) +
+  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
+  coord_polar("y", start = 0) +
+  geom_text(aes(label = paste0(percentage, "%")), 
+            position = position_stack(vjust = 0.5), 
+            size = 6,   # Increase text size for percentage labels
+            fontface = "bold") +  # Make percentage text bold
+  scale_fill_manual(values = upgrade_colors_low) +  # Apply colors for Upgraded vs Not Upgraded
+  labs(fill = "Upgrade Status") +
+  theme_void() +  # Remove axis, grid lines, and background
+  theme(legend.position = "right", 
+        legend.title = element_text(size = 16),  # Increase legend title size
+        legend.text = element_text(size = 14))  # Increase legend text size
+
+# Filter data for 'Moderate Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
+moderate_comfort_data <- lane_type_data %>%
+  filter(Comfort_Level == "Moderate Comfort") %>%
+  mutate(
+    Upgrade_Status = ifelse(is.na(UPGRADED), "Not Upgraded", "Upgraded")
+  ) %>%
+  group_by(Upgrade_Status) %>%
+  summarise(count = n())  # Count the occurrences of each category
+
+# Calculate percentages
+moderate_comfort_data <- moderate_comfort_data %>%
+  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column
+
+# Define colors for 'Upgraded' and 'Not Upgraded' categories
+upgrade_colors_moderate <- c(
+  "Upgraded" = "#8A7967",  # Gray-brown for upgraded
+  "Not Upgraded" = "#B39B8E"  # Muted pinkish tone for not upgraded
+)
+
+# Generate the pie chart for Moderate Comfort lanes
+ggplot(moderate_comfort_data, aes(x = "", y = percentage, fill = Upgrade_Status)) +
+  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
+  coord_polar("y", start = 0) +
+  geom_text(aes(label = paste0(percentage, "%")), 
+            position = position_stack(vjust = 0.5), 
+            size = 6,   # Increase text size for percentage labels
+            fontface = "bold") +  # Make percentage text bold
+  scale_fill_manual(values = upgrade_colors_moderate) +  # Apply colors for Upgraded vs Not Upgraded
+  labs(fill = "Upgrade Status") +
+  theme_void() +  # Remove axis, grid lines, and background
+  theme(legend.position = "right", 
+        legend.title = element_text(size = 16),  # Increase legend title size
+        legend.text = element_text(size = 14))  # Increase legend text size
+
+# Filter data for 'High Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
+high_comfort_data <- lane_type_data %>%
+  filter(Comfort_Level == "High Comfort") %>%
+  mutate(
+    Upgrade_Status = ifelse(is.na(UPGRADED), "Not Upgraded", "Upgraded")
+  ) %>%
+  group_by(Upgrade_Status) %>%
+  summarise(count = n())  # Count the occurrences of each category
+
+# Calculate percentages
+high_comfort_data <- high_comfort_data %>%
+  mutate(percentage = round((count / sum(count)) * 100, 2))  # Add percentage column
+
+# Define colors for 'Upgraded' and 'Not Upgraded' categories
+upgrade_colors_high <- c(
+  "Upgraded" = "#8A7967",  # Gray-brown for upgraded
+  "Not Upgraded" = "#B39B8E"  # Muted pinkish tone for not upgraded
+)
+
+# Generate the pie chart for High Comfort lanes
+ggplot(high_comfort_data, aes(x = "", y = percentage, fill = Upgrade_Status)) +
+  geom_bar(stat = "identity", width = 1, color = "black") +  # Add black border lines around pie sections
+  coord_polar("y", start = 0) +
+  geom_text(aes(label = paste0(percentage, "%")), 
+            position = position_stack(vjust = 0.5), 
+            size = 6,   # Increase text size for percentage labels
+            fontface = "bold") +  # Make percentage text bold
+  scale_fill_manual(values = upgrade_colors_high) +  # Apply colors for Upgraded vs Not Upgraded
+  labs(fill = "Upgrade Status") +
+  theme_void() +  # Remove axis, grid lines, and background
+  theme(legend.position = "right", 
+        legend.title = element_text(size = 16),  # Increase legend title size
+        legend.text = element_text(size = 14))  # Increase legend text size
+
+```
+
 ## Maps
 
-Our results are summarized in @tbl-modelresults.
+```{r}
+#| label: fig-map
+#| fig-cap: "Map of Bike Lanes Locations in Toronto"
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-width: 11
+#| fig-height: 8
+
+# Load the data
+coordinates_data <- read_csv(here("data", "analysis_data", "coordinates_data.csv"))
+
+# Get bounding box for Toronto (approximate limits for Toronto)
+toronto_bbox <- c(-79.6303856025883, 43.5822069220627, -79.1180336515019, 43.8554571861712)
+
+# Fetch map data from OpenStreetMap (OSM)
+toronto_map <- opq(bbox = toronto_bbox) %>%
+  add_osm_feature(key = 'highway', value = 'primary') %>%
+  osmdata_sf()
+
+# Convert map data to Simple Features (sf) objects
+streets <- toronto_map$osm_lines
+
+# Plot the map with bike lanes data
+ggplot() +
+  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) +  # OSM street map layer
+  geom_point(data = coordinates_data, aes(x = longitude, y = latitude), 
+             size = 0.1, color = "black") +  # Bike lanes points
+  labs(x = "Longitude", y = "Latitude") +
+  coord_sf(xlim = c(-79.6303856025883, -79.1180336515019), 
+           ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE) +
+  theme_minimal() +
+  theme(
+    panel.background = element_blank(),  # Remove panel background
+    panel.grid.major = element_blank(),  # Remove major grid lines
+    panel.grid.minor = element_blank(),  # Remove minor grid lines
+    axis.text.x = element_text(angle = 45, hjust = 1)
+  )
+
+```
+
+@fig-map provides a visual representation of the bike lane infrastructure across Toronto.
+The lines on the map correspond to various cycling paths, including dedicated bike lanes, multi-use paths, and shared roadways, which span different areas of the city.
+The geographic coordinates, ranging from around 43.60°N to 43.85°N in latitude and -79.6°W to -79.3°W in longitude, encompass most of Toronto’s urban region.
+This map offers an important spatial perspective, illustrating the extensive network of cycling routes that connect different neighborhoods and areas within the city.
+
+A key observation from the map is the dense concentration of bike lanes in the downtown core and along major roadways, indicating a well-established network in central Toronto.
+The downtown area’s grid-like structure is particularly evident, showcasing a high density of lanes that facilitate easy cycling access within the city's busiest areas.
+As one moves further away from the city center, the map shows fewer bike lanes, suggesting that suburban regions may have less developed cycling infrastructure.
+However, there are notable multi-use paths extending toward the outskirts, particularly in parks and recreational areas, providing important cycling connections for those commuting or using bikes for recreational purposes.
+
+This map highlights the importance of cycling infrastructure in promoting sustainable urban transportation.
+The wide distribution of bike lanes supports Toronto's efforts to enhance active transportation options, reduce traffic congestion, and improve accessibility for cyclists.
+The representation also provides a foundation for analyzing areas where future infrastructure developments could be prioritized, particularly in regions where cycling connectivity appears sparse.
 

### 2024-09-24 paper/paper.qmd
paper results section updated

@@
   kable("latex", booktabs = TRUE, col.names = c("Simplified Geometry")) %>%
-  kable_styling(latex_options = c("hold_position")) %>%
   column_spec(1, width = "40em")  # Adjust the width as needed
@@
 
+@fig-installed consists of two panels depicting the number of bikeways installed in Toronto by year, alongside a trend analysis of the installation pattern over time.
+@fig-installed-1 displays the raw count of bikeways installed per year from 2001 to 2024.
+A significant spike is observed in 2001, with over 400 bikeways installed in that year alone.
+This likely reflects a major investment in cycling infrastructure at the time.
+Following 2001, the number of installations drops sharply and stabilizes at much lower levels in subsequent years, with occasional small peaks around 2007, 2011, and 2021.
+
+@fig-installed-2 adds a trend line to the same data, showing a gradual decline in the number of bikeways installed over time.
+The trend line, sloping downward, indicates that despite some fluctuations, the overall pattern is a decrease in annual bikeway installations.
+The early 2000s marked a high point for cycling infrastructure development, but after this surge, the rate of installation has steadily declined.
+This trend might reflect a shift in city planning priorities or the saturation of available spaces for bikeways in certain areas.
+The analysis highlights the need for renewed focus on expanding cycling infrastructure to sustain or increase the momentum of bike-friendly urban development.
+
 ```{r}
@@
 
+@fig-upgraded illustrates the number of bikeways upgraded in Toronto from 2001 to 2023, with panel (a) showing the raw counts of upgrades and panel (b) visualizing the trend.
+@fig-upgraded-1 reveals that the most significant number of upgrades occurred in 2015, where nearly 100 bikeways were updated, marking a major peak.
+Outside of this spike, the number of annual upgrades remains relatively modest, typically fluctuating between 10 and 50 bikeways.
+Notable periods of increased upgrades are visible around 2012 and 2020, possibly reflecting concentrated efforts to modernize or expand existing cycling infrastructure in those years.
+
+In @fig-upgraded-2, the overall trend is slightly upward, as indicated by the red trend line.
+Despite the notable peaks and valleys, the trend line suggests that, on average, the number of bikeway upgrades has increased over the years.
+The spike in 2015 significantly affects the trend, but even excluding that, the gradual rise suggests that more consistent efforts have been made to maintain and improve Toronto’s cycling network.
+This trend aligns with broader urban planning strategies focused on making cities more accessible and environmentally friendly by supporting active transportation.
+The dips in recent years, particularly after 2020, might be influenced by resource reallocation or external factors, such as the global pandemic, which could have delayed infrastructure projects.
+
 ## Distributions
@@
 ```{r}
-#| label: fig-comfort
+#| label: fig-comfort-one
 #| fig-cap: "Toronto Bikeway Comfort Level Distribution"
-#| fig-subcap: ["All levels", "Low Comfort", "Moderate Comfort", "High Comfort"]
+#| fig-subcap: ["Comfort Levels", "Low Comfort"]
 #| echo: false
@@
 #| layout-ncol: 2
-#| layout-nrow: 2
+#| layout-nrow: 1
+#| fig-width: 12
+#| fig-height: 7
 
@@
 
+```
+
+@fig-comfort-one provides a breakdown of Toronto’s bikeway network by comfort level and upgrade status.
+@fig-comfort-one-1 shows the distribution of all bikeways based on comfort level, with the majority of bikeways classified as either Low Comfort (41.6%) or Moderate Comfort (41.1%), while High Comfort lanes constitute a smaller portion (17.3%).
+This indicates that while Toronto has invested in a significant number of bike lanes, most of them fall within lower comfort categories, suggesting that further improvements, particularly in upgrading existing lanes, are needed to promote safer and more comfortable cycling experiences.
+
+```{r}
+#| label: fig-comfort-two
+#| fig-cap: "Toronto Bikeway Comfort Level Distribution"
+#| fig-subcap: ["Moderate Comfort", "High Comfort"]
+#| echo: false
+#| warning: false
+#| message: false
+#| layout-ncol: 2
+#| layout-nrow: 1
+#| fig-width: 12
+#| fig-height: 7
+
+# Load the data
+lane_type_data <- read_csv(here("data", "analysis_data", "lane_type_data.csv"))
+
 # Filter data for 'Moderate Comfort' lanes and categorize as 'Upgraded' or 'Not Upgraded'
@@
 ```
+
+@fig-comfort-one-2, @fig-comfort-two-1, and @fig-comfort-two-2 offer a deeper look into the upgrade status for each comfort level.
+For Low Comfort lanes, only 22.69% have been upgraded, while 77.31% remain in their original state.
+Similarly, for Moderate Comfort bikeways, 23.38% have been upgraded, leaving 76.62% untouched.
+Interestingly, High Comfort lanes have the lowest proportion of upgrades, with only 18.76% improved, despite representing the most comfortable and potentially most desirable biking infrastructure.
+These trends highlight a potential gap in Toronto's focus on maintaining and upgrading its most comfortable and accessible cycling lanes, suggesting a need for more targeted investments in high-comfort bikeways to encourage broader use of cycling for transportation.
 

### 2024-09-24 paper/paper.qmd
paper discussion section added

@@
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+The results highlight Toronto's phased approach to developing its cycling infrastructure, with a significant spike in installations in the early 2000s, as shown in @fig-installed.
+This surge is likely because the dataset started recording from 2001, hence all installations before was counted towards 2001.
+Futher, the fact that goverment start taking this record and the large number of installations at this period may have reflected political will and public demand for greener transportation alternatives, driven by growing concerns about climate change and urban congestion.
+
+However, after the initial surge, the number of bikeways installed annually began to decline, a pattern that persists to this day.
+Several factors could explain this decline.
+First, the initial phase of the bike plan likely focused on areas where cycling infrastructure could be installed with relative ease, particularly in downtown Toronto, where road networks and urban density supported such investments.
+As the network expanded into more complex areas, such as suburban neighborhoods with fewer existing roadways, the pace of installation slowed.
+Additionally, the physical limitations of available space in a highly developed urban area may have constrained further expansion, as new bikeways would require more complicated and costly measures, such as road reallocation or shared-use paths.
+
+Despite this slowdown in new installations, @fig-upgraded demonstrates a significant focus on upgrading existing infrastructure, particularly in later years.
+This shift from installation to upgrades could indicate a strategic pivot by the city toward improving the quality and safety of existing cycling infrastructure.
+As infrastructure ages and new safety standards emerge, upgrading older lanes becomes crucial to ensuring the network remains usable and safe for cyclists.
+The peak of upgrades in 2015, for example, might reflect the city's response to increased cycling activity, demands from advocacy groups, or the implementation of updated cycling policies that prioritize safer and more modern designs, such as protected bike lanes.
+
+This pattern of prioritizing upgrades over new installations also suggests that Toronto's initial expansion may have left some areas with suboptimal or unsafe infrastructure, particularly in the form of shared roadways or painted bike lanes that offer minimal protection to cyclists.
+As urban transportation planning evolves, the focus has shifted toward creating more separated and protected lanes, which are essential for encouraging cycling among less experienced riders, families, or those concerned about safety.
+This ongoing commitment to improving the infrastructure—rather than merely expanding it—reflects a maturation in Toronto’s approach to active transportation, as planners recognize that quality and safety are just as important as quantity when promoting cycling as a viable mode of transport.
+
+Finally, another crucial aspect linked to the decline in installations could be budgetary constraints or competing infrastructure projects.
+As urban budgets are stretched across various priorities—public transit expansion, road maintenance, and green space development—the pace of installing new cycling lanes might slow.
+However, the investment in upgrades suggests that cycling infrastructure remains a priority, even if the focus has shifted from expansion to refinement.
+Continuing to monitor and analyze these trends will be essential for understanding how Toronto can balance its need for both new installations and the improvement of existing cycling routes to meet its long-term sustainability and mobility objectives.
+
+In conclusion, the first discussion point highlights the evolution of Toronto’s cycling infrastructure from an initial phase of rapid expansion to a more recent focus on upgrading and enhancing the safety and quality of the network.
+This shift reflects the challenges of balancing growth with maintenance, the need to adapt to changing urban landscapes, and the influence of global best practices and local advocacy in shaping the city’s cycling future.
 
@@
 
+The comfort level classification of Toronto's cycling infrastructure offers valuable insights into the quality of the city's bike network, particularly with respect to cyclist safety and usability.
+As shown in @fig-comfort-one, the majority of bikeways fall into the Low or High Comfort categories, with Moderate Comfort lanes—typically those offering relatively protected infrastructure—making up a smaller proportion of the overall network.
+This distribution suggests that, Toronto has successfully expanded the sheer number of bikeways, the majority could offer the protection and comfort necessary to encourage broader and more consistent usage.
+However, it is better to see Low Comfort level lanes taking the smallest portion among all.
+
+The classification of cycling lanes into different comfort levels is important because it directly affects who feels confident enough to cycle.
+Low Comfort lanes, often painted bike lanes or shared roadways, offer minimal physical separation from motor vehicles, making them less appealing to risk-averse cyclists.
+The lack of protection increases the perceived danger, discouraging people from cycling regularly, particularly in areas with high traffic volumes.
+For example, a cyclist may feel comfortable using a Low Comfort lane in a quiet residential neighborhood but avoid it entirely on a busy downtown street where traffic and intersections pose greater hazards.
+
+Moderate Comfort lanes, which typically include designated bike lanes, offer more safety but still lack the physical separation provided by protected or High Comfort infrastructure.
+These lanes are a step in the right direction but may not sufficiently address concerns about safety, particularly in areas with fast-moving or congested vehicle traffic.
+Moderate Comfort lanes are likely to attract more cyclists than Low Comfort lanes, but they may not be enough to significantly increase ridership across all demographic groups.
+For instance, studies have shown that women, children, and older adults are particularly sensitive to perceived danger in cycling, and without sufficient protection, these groups are less likely to use the infrastructure regularly.
+
+High Comfort lanes, on the other hand, provide the greatest potential for encouraging cycling across a broad range of demographics.
+These lanes, often in the form of protected bike lanes or multi-use paths, offer a physically separated space for cyclists, reducing the likelihood of accidents involving motor vehicles.
+Research from cities that have invested heavily in High Comfort infrastructure, such as Amsterdam or Copenhagen, indicates that the availability of such protected lanes is a key factor in increasing overall cycling rates and making cycling a more viable and appealing transportation option for all.
+In Toronto, however, as the analysis shows, only a small proportion of the network is classified as High Comfort.
+This limits the extent to which the city can fully leverage its cycling infrastructure to promote widespread, sustained usage.
+
+The low proportion of High Comfort lanes can be partly attributed to the early phases of Toronto's cycling infrastructure development, during which the focus was likely on quickly installing a large number of lanes to meet political and public demands.
+At that time, lower-cost solutions like painted bike lanes and sharrows (shared roadways) were easier and faster to implement than more expensive, space-consuming, and structurally complex protected lanes.
+However, these lanes did not necessarily provide the level of safety needed to support a significant increase in cycling activity, particularly among more cautious cyclists.
+The subsequent shift towards upgrading infrastructure, as discussed earlier, is likely an acknowledgment of the limitations of the initial approach, where quantity may have been prioritized over quality.
+
+Moreover, the analysis reveals that many of Toronto's Moderate and Low Comfort lanes have yet to be upgraded, as shown in @fig-comfort-one-2 and @fig-comfort-two.
+For both Low and Moderate Comfort lanes, over 75% have not been upgraded, which implies that much of the city's existing cycling network still operates under outdated or less protective standards.
+This points to a significant gap between the current state of Toronto's cycling infrastructure and the standards that are increasingly being recognized as necessary to make cycling a mainstream mode of transportation in urban environments.
+The failure to upgrade a larger portion of these lanes may stem from budgetary constraints, political will, or challenges in reallocating road space in already congested or car-dominated areas.
+
+One of the most critical implications of these findings is that Toronto’s cycling infrastructure might not yet be sufficiently robust to meet the city’s broader goals of reducing car dependency, lowering emissions, and promoting healthier, more active lifestyles.
+To truly shift a significant portion of the population towards cycling, the city will need to invest more heavily in upgrading its infrastructure, particularly in transitioning more of its Low and Moderate Comfort lanes into High Comfort lanes.
+This is especially important in areas where traffic volume and speed are higher, as these conditions pose greater risks to cyclists and deter people from choosing bikes as a mode of transportation.
+
+Another potential reason for the focus on Moderate and Low Comfort lanes is the spatial distribution of the city’s bikeways.
+As mentioned earlier when showing the result of @fog-map, much of Toronto’s cycling network is concentrated in the downtown core, where road space is at a premium.
+Installing protected lanes in these areas often requires taking road space away from cars, which can be politically and technically challenging.
+In contrast, suburban areas, where road space is more abundant, could offer better opportunities for the city to build High Comfort infrastructure.
+Expanding the reach of the cycling network into these suburban areas, while upgrading the comfort levels of existing downtown lanes, could have a transformative effect on overall cycling rates and help make cycling a viable option for longer commutes.
+
+In conclusion, while Toronto has made considerable progress in expanding its cycling infrastructure, the comfort level analysis reveals that much of the network remains suboptimal for ensuring safety and encouraging more widespread use.
+To achieve its sustainability and transportation goals, the city must focus on upgrading its existing Low and Moderate Comfort lanes and prioritize the construction of High Comfort infrastructure.
+This shift will be crucial not only for improving safety but also for making cycling an attractive option for a more diverse range of residents, thereby promoting equity and inclusivity in urban transportation.
+
 ## Third discussion point
+
+The geospatial analysis of Toronto's cycling infrastructure, as visualized in @fig-map, reveals clear spatial patterns that highlight both the strengths and weaknesses of the city’s current bike network.
+One of the key findings is the dense concentration of bikeways in the downtown core, where cycling infrastructure is most integrated into the existing urban fabric.
+This concentration aligns with expectations, as downtown Toronto is the city’s most densely populated and heavily trafficked area, where the demand for alternative transportation modes is highest.
+The high density of bikeways in the city center reflects a concerted effort to provide accessible cycling options in areas where they are most likely to be used, particularly by commuters and urban residents.
+
+However, the map also exposes a noticeable gap in cycling infrastructure as one moves further from the city center.
+The relative sparsity of bike lanes in suburban and peripheral areas suggests that while downtown residents enjoy a well-connected cycling network, those living in the suburbs have fewer options for safe, comfortable bike routes.
+This discrepancy may limit the overall effectiveness of Toronto’s cycling infrastructure, as it primarily serves those in the urban core while potentially neglecting a significant portion of the population.
+Suburban residents, who often have longer commutes and fewer public transit options, could benefit greatly from expanded cycling infrastructure that offers a viable alternative to car travel.
+
+The lack of infrastructure in suburban areas could be attributed to a variety of factors.
+Suburbs typically have lower population densities and wider roads, making cycling infrastructure less of a priority in urban planning compared to the densely populated downtown core.
+Additionally, suburban regions tend to have more car-dependent cultures, where cycling is not as ingrained as a common mode of transportation.
+The existing infrastructure may therefore reflect historical priorities that favored cars over active transportation.
+However, as cities like Toronto strive to reduce traffic congestion and meet climate goals, expanding cycling infrastructure into these suburban areas could be a crucial step toward encouraging more sustainable transportation choices.
+
+Furthermore, the map shows the importance of major recreational areas, such as parks and waterfronts, in Toronto’s cycling network.
+Multi-use trails, which provide High Comfort lanes, are particularly prevalent in these areas, offering scenic and safe routes for cyclists.
+While these trails are beneficial for recreational cycling, they may not be fully integrated into the city’s transportation network, limiting their utility for commuters.
+Connecting these trails to the downtown core and other major employment hubs could improve their role in promoting cycling as a legitimate form of daily transportation rather than just a leisure activity.
+
+Lastly, the map highlights potential future opportunities for infrastructure expansion.
+Areas with sparse cycling routes could be prioritized for new installations, especially where there is existing road capacity or public demand for safer cycling options.
+Identifying and addressing these gaps will be crucial as Toronto continues to grow and seeks to make its transportation system more equitable and sustainable.
+The current infrastructure is heavily concentrated in downtown areas, but future developments could focus on creating a more geographically balanced network, ensuring that all residents, regardless of location, have access to safe and convenient cycling routes.
+
+In summary, the geospatial distribution of Toronto’s cycling infrastructure reveals a well-connected downtown network but highlights the need for more investment in suburban areas and better integration of recreational trails into the broader transportation system.
+Expanding the reach of cycling infrastructure into underdeveloped areas will be key to making cycling a viable transportation option for a larger portion of the city’s population.
 
@@
 
-Weaknesses and next steps should also be included.
+One key limitation of this study is the reliance on the 'Cycling Network' dataset from OpenDataToronto, which, while comprehensive, might not capture informal or emerging cycling routes that are not officially recognized.
+Additionally, the dataset may not include newer infrastructure developments if updates were delayed in publication.
+Future research should integrate real-time data collection techniques or alternative datasets (e.g., crowdsourced information) to ensure that the analysis reflects the most current state of the network.
+
+Another weakness lies in the classification of comfort levels, which, while useful, is a simplification.
+Comfort in cycling can be influenced by numerous factors beyond infrastructure type, such as traffic density, road conditions, and seasonal weather.
+A more nuanced classification that incorporates these additional factors could provide a more accurate assessment of how comfortable Toronto's cycling network truly is for different types of cyclists.
+
+Next steps should include focusing on identifying areas with fewer bike lanes and lower comfort levels, particularly in suburban areas.
+These regions could be targeted for infrastructure upgrades and expansion to promote greater equity in access to safe cycling options.
+Additionally, analyzing cycling usage patterns, perhaps through mobile GPS data or surveys, would offer valuable insights into how the infrastructure is being utilized, helping city planners to make data-driven decisions in future expansions or upgrades.
 

### 2024-09-24 paper/paper.qmd
paper titles, abstruct, intro added

@@
 ---
-title: "My title"
-subtitle: "My subtitle"
+title: "Geospatial and Temporal Analysis of Toronto's Cycling Infrastructure"
+subtitle: "A Data-Driven Approach"
 author: 
@@
 date-format: long
-abstract: "My abstract"
+abstract: "This paper explores the development of Toronto's cycling infrastructure.
+Through a comprehensive analysis of installation dates, upgrades, and comfort levels of bike lanes, this study highlights the trends in cycling infrastructure expansion from 2001 to 2024.
+A significant focus is placed on the comfort levels of different types of lanes, offering insights into how the infrastructure meets the needs of cyclists.
+The spatial distribution of bike lanes is also analyzed, revealing discrepancies in access between downtown and suburban areas.
+The findings emphasize the need for continued investment in both the quality and expansion of cycling infrastructure to promote sustainable transportation in Toronto."
 format: pdf
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR.
+Urban cycling has become an increasingly vital component of sustainable transportation strategies in many cities, including Toronto.
+As the city grows and urban congestion intensifies, promoting active transportation options such as cycling is essential for reducing emissions, improving public health, and enhancing mobility.
+Toronto's efforts to develop a comprehensive cycling infrastructure began in the early 2000s, resulting in a network of bike lanes, multi-use trails, and shared roadways designed to support cyclists across the city.
 
-The remainder of this paper is structured as follows.
+This paper aims to analyze the trends in Toronto’s cycling infrastructure, focusing on the patterns of bikeway installations, upgrades, and the classification of bike lanes by comfort level.
+Using data from the 'Cycling Network' dataset provided by Open Data Toronto, which will be discussed in detail at @sec-data.
+The study covers the evolution of cycling infrastructure over the past two decades, highlighting both progress and challenges.
+Additionally, the geographic distribution of bike lanes across the city is examined to identify areas with potential for further development.
+
+The remainder of this paper is structured as follows: the data sources and methodologies are presented in @sec-data, followed by an analysis of the results in @sec-res, and a discussion of the findings in @sec-dis.
 

### 2024-09-24 paper/paper.qmd
paper references updated

@@
 Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR].
-Also with support of additional packages in R: `sf` [], `readr` [@citeReadr], `ggplot2` [@citeGgplot], `osmdata` [], `tidyverse` [@citeTidyverse], `jsonlite` [], `dplyr` [@citeDplyr], `here` [], `knitr` [], `kableExtra` [], `` [], `` [], .
+Also with support of additional packages in R: `sf` [@citeSf], `readr` [@citeReadr], `ggplot2` [@citeGgplot2], `osmdata` [@citeOsmdata], `tidyverse` [@citeTidyverse], `jsonlite` [@citeJsonlite], `dplyr` [@citeDplyr], `here` [@citeHere], `knitr` [@citeKnitr], `kableExtra` [@citeKableExtra].
 

### 2024-09-24 paper/paper.qmd
paper data section updated, appendix updated

@@
 
-Urban cycling has become an increasingly vital component of sustainable transportation strategies in many cities, including Toronto.
-As the city grows and urban congestion intensifies, promoting active transportation options such as cycling is essential for reducing emissions, improving public health, and enhancing mobility.
-Toronto's efforts to develop a comprehensive cycling infrastructure began in the early 2000s, resulting in a network of bike lanes, multi-use trails, and shared roadways designed to support cyclists across the city.
+Urban cycling plays a critical role in advancing sustainable transportation strategies, especially in rapidly growing cities like Toronto.
+With the city facing increasing congestion, promoting active transportation modes such as cycling is essential for reducing greenhouse gas emissions, improving public health, and enhancing urban mobility.
+Since the early 2000s, Toronto has made significant strides in expanding its cycling infrastructure, creating a network of bike lanes, multi-use paths, and shared roadways to support cyclists citywide.
 
-This paper aims to analyze the trends in Toronto’s cycling infrastructure, focusing on the patterns of bikeway installations, upgrades, and the classification of bike lanes by comfort level.
-Using data from the 'Cycling Network' dataset provided by Open Data Toronto, which will be discussed in detail at @sec-data.
-The study covers the evolution of cycling infrastructure over the past two decades, highlighting both progress and challenges.
-Additionally, the geographic distribution of bike lanes across the city is examined to identify areas with potential for further development.
+This paper analyzes the evolution of Toronto's cycling infrastructure over the past two decades, focusing on the installation and upgrade trends of bikeways and their classification by comfort level.
+Using the 'Cycling Network' dataset from Open Data Toronto, the study examines both the spatial distribution of bike lanes and the varying levels of cycling comfort across different regions.
+The analysis highlights discrepancies between downtown and suburban areas, showcasing progress while identifying regions that remain underserved.
 
-The remainder of this paper is structured as follows: the data sources and methodologies are presented in @sec-data, followed by an analysis of the results in @sec-res, and a discussion of the findings in @sec-dis.
+The study's findings underscore the need for continued investment in both the expansion and quality of Toronto's cycling infrastructure to encourage more widespread adoption of cycling as a viable mode of transportation.
+By addressing both the geographic imbalances and comfort levels of bike lanes, the city can better meet its sustainability goals.
+
+The structure of this paper is as follows: the data sources and methodologies are presented in @sec-data, followed by an analysis of the results in @sec-res, and a discussion of the findings in @sec-dis.
 
@@
 ## Source
+The dataset analyzed in this study, the Cycling Network dataset [@citeCyclingNetwork], was retrieved from Open Data Toronto [@citeODT].
+This dataset is publicly available and is continually updated to reflect Toronto’s ongoing efforts to develop and maintain its cycling infrastructure.
+It contains geospatial and temporal information about various cycling-related features, including the types of bikeways (dedicated lanes, shared roads, multi-use trails), installation years, and any subsequent upgrades.
+By offering a detailed history of Toronto's cycling network development, this dataset plays a vital role in informing both urban planning policies and research into sustainable transportation options.
 
-The data used in this paper was collected by the OpenDataToronto Library [@citeODT].
-OpenDataToronto provides a platform for the public to access various datasets related to Toronto’s civic operations and urban infrastructure.
-The specific dataset used in this research is the 'Cycling Network' [@citeDataset], which provides detailed information on Toronto’s bicycle infrastructure, including dedicated lanes, multi-use trails, and shared roadways.
-This dataset plays a crucial role in promoting active transportation and urban sustainability efforts within the city.
-It is frequently updated to reflect ongoing expansions and modifications to the cycling infrastructure, aligning with Toronto's broader efforts to reduce congestion, enhance mobility, and support environmental goals.
-This dataset not only informs local decisions but also contributes to broader discussions on urban transportation planning and sustainability in growing cities around the world.
+While the Cycling Network dataset was the primary source of data, other relevant datasets were explored, such as the Road Network File [@citeStatCan] from Statistics Canada, which provides detailed geospatial data on all roadways in Toronto.
+Another such alternative was the Major City-Wide Cycling Routes dataset [@citeCyclingRoutes], also available on Open Data Toronto [@citeODT].
+However, this dataset focuses specifically on the large-scale cycling routes connecting different parts of the city, emphasizing connectivity at a macro level.
+While these two dataset are useful for analyzing high-level transportation patterns, it lacks the detailed, segment-level data—such as comfort levels, installation dates, and upgrade history—that are crucial for a deep analysis of cycling infrastructure.
+Since this paper focuses on both the growth and quality of cycling infrastructure, the Cycling Network dataset, with its more comprehensive coverage of individual bikeway segments, was a better fit for the goals of this study.
 
-Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages which will be talked about in the @sec-data-method.
+Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages which will be mentioned later in the @sec-data-method.
 
@@
 This allows researchers to analyze trends in cycling infrastructure and its expansion over time.
+
+### Variables
+
+As mentioend above, the dataset contains several key variables that were used in this analysis:
+
+#### Installation Year
+
+The Installation Year records the year in which a particular bikeway segment was first constructed or made available for use.
+This variable is essential for conducting a temporal trend analysis, allowing the study to track the growth and expansion of the cycling network over time.
+
+#### Upgrade Year
+
+The Upgrade Year indicates when an existing bikeway segment was upgraded or improved.
+Upgrades typically involve safety enhancements, such as the installation of protective barriers, widening of lanes, or changes from painted lanes to physically separated lanes.
+This variable is critical for assessing not just the expansion but also the quality improvements made to Toronto’s cycling infrastructure over time.
+Tracking upgrades helps to understand whether the city's focus has shifted from merely adding new bikeways to improving the usability and safety of existing ones, especially in high-traffic or critical areas.
+
+#### Bikeway Type
+
+The Bikeway Type categorizes the cycling infrastructure into different types, including dedicated bike lanes, multi-use trails, and shared roadways, etc.
+Each type of bikeway offers varying levels of protection and comfort to cyclists.
+This variable was not directly used in the study, but forming another crucial variable: Comfort Level.
+
+#### Comfort Level
+Each bikeway is assigned a comfort level—high, moderate, or low—based according to the recorded Bikeway Type.
+These classification are self-defined, reling on factors such as physical separation from traffic and lane width of each Bikeway Type, detail can be found in @sec-app-comfort.
+This variable is central to understanding the usability of the infrastructure.
+
+#### Geospatial Coordinates
+Latitude and longitude data are included to map the bikeways and assess their spatial distribution.
 
@@
 head(data_geometry) %>%
-  kable("latex", booktabs = TRUE, col.names = c("Simplified Geometry")) %>%
+  kable("latex", booktabs = TRUE, col.names = c("Geometry")) %>%
   column_spec(1, width = "40em")  # Adjust the width as needed
@@
 Data from the 'INFRA_HIGHORDER' and 'INFRA_LOWORDER' columns was cleaned, and missing values were removed.
-The lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, depending on whether the infrastructure was protected or involved bike lanes.
+The lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, depending on their Bikeway Type data, detail of this classification can be found in @sec-app-comfort.
 The classification, along with year data for installation and upgrades, was saved for analysis on the types of cycling lanes and their comfort levels.
@@
 
+## Comfort Levels Classification {#sec-app-comfort}
+
+### Data
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-comfort
+#| tbl-cap: "INFRA_HIGHORDER and INFRA_LOWORDER Columns from Raw Data"
+#| warning: false
+
+# Load the dataset
+data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+
+comfort_data <- data %>%
+  select(INFRA_HIGHORDER, INFRA_LOWORDER)
+
+head(comfort_data) %>%
+  kable("latex", booktabs = TRUE, col.names = c("INFRA_HIGHORDER", "INFRA_LOWORDER"))
+
+```
+
+The classification relies on the infrastructure types defined in two key columns: INFRA_HIGHORDER and INFRA_LOWORDER, shown in @tbl-comfort.
+As the author explained on the original dataset page [@citeCyclingNetwork]: "'INFRA-HIGH' refers to the highest classified cycling infrastructure type installed on this segment.
+'INFRA-LOW' refers to the lowest classified cycling infrastructure type.
+The hierarchy is based on the quality and comfort of the bikeway type, with routes that are shared with vehicles as the lowest comfort, and the greater the separation / dedicated space, the higher the comfort and classification."
+
+Both columns are crucial because they give complementary information about the physical features of the bikeway.
+By selecting these columns and ensuring that no missing values remain, the dataset is prepared for classification.
+
+### Logic
+
+The comfort levels are defined using a custom classification scheme based on the infrastructure types.
+The classification assigns a comfort level to each bikeway segment depending on its physical characteristics and separation from traffic:
+
+#### High Comfort
+Segments with infrastructure types labeled as "Protected" or "Multi-Use." These bikeways generally have physical barriers separating cyclists from motor vehicles, such as fully separated bike lanes or paths shared with pedestrians but removed from vehicle traffic.
+This category offers the highest level of safety and comfort, making it appealing to a wide range of cyclists, including families and less experienced riders.
+
+#### Moderate Comfort
+Bikeways labeled as "Bike Lane" in either column are classified as "Moderate Comfort." These lanes are typically painted or designated lanes on the road but lack the full physical separation that would offer more safety.
+While these lanes provide cyclists with their own space, they are adjacent to motor vehicle traffic, which may make them less comfortable for some users.
+
+#### Low Comfort
+All other types of bikeways, not falling into the protected or designated bike lane categories, are classified as "Low Comfort." These may include shared roadways where cyclists must navigate mixed traffic, offering little or no physical protection from vehicles.
+These segments are considered the least comfortable and safest for cyclists, often deterring usage by more risk-averse cyclists.
+
+This classification method ensures that each bikeway segment is categorized into one of the three comfort levels based on its infrastructure type, enabling a detailed analysis of the usability of Toronto's cycling network.
+
 \newpage

### 2024-09-24 paper/paper.qmd
paper first draft

@@
 #| label: tbl-clean-upgraded
-#| tbl-cap: "Number of Lane Upgrades by Year"
+#| tbl-cap: "Number of Bikeways Upgraded by Year"
 #| warning: false
@@
 tail(upgraded_summary) %>%
-  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Upgraded Lanes"))
+  kable("latex", booktabs = TRUE, col.names = c("Year", "Total Upgraded"))
 
@@
 head(installed_summary) %>%
-  kable("latex", booktabs = TRUE, col.names = c("Year", "Number of Installed Bikeways"))
+  kable("latex", booktabs = TRUE, col.names = c("Year", "Total Installed"))
 
@@
 
-## First discussion point
+## Development of Toronto's Cycling Infrastructure
 
-The results highlight Toronto's phased approach to developing its cycling infrastructure, with a significant spike in installations in the early 2000s, as shown in @fig-installed.
-This surge is likely because the dataset started recording from 2001, hence all installations before was counted towards 2001.
-Futher, the fact that goverment start taking this record and the large number of installations at this period may have reflected political will and public demand for greener transportation alternatives, driven by growing concerns about climate change and urban congestion.
+The data reveals a phased development of Toronto's cycling infrastructure, with a notable increase in installations during the early 2000s, as shown in @fig-installed.
+This rise is likely due to the dataset commencing in 2001, thereby consolidating all installations prior to this year.
+Additionally, the onset of formal record-keeping and the significant number of installations during this period may reflect political momentum and public demand for environmentally sustainable transportation alternatives, influenced by growing concerns over climate change and urban congestion.
 
-However, after the initial surge, the number of bikeways installed annually began to decline, a pattern that persists to this day.
-Several factors could explain this decline.
-First, the initial phase of the bike plan likely focused on areas where cycling infrastructure could be installed with relative ease, particularly in downtown Toronto, where road networks and urban density supported such investments.
-As the network expanded into more complex areas, such as suburban neighborhoods with fewer existing roadways, the pace of installation slowed.
-Additionally, the physical limitations of available space in a highly developed urban area may have constrained further expansion, as new bikeways would require more complicated and costly measures, such as road reallocation or shared-use paths.
+Ignoring the data from the first year, the annual number of new bikeway installations still showing a trend of slight decreasing.
+Several factors may account for this decline.
+In the early stages, the city's cycling plan likely targeted areas where infrastructure could be installed with relative ease, particularly in downtown Toronto, where road networks and urban density favored such investments.
+As the network extended into more complex regions, such as suburban areas with fewer existing roadways, the rate of installation slowed.
+Moreover, the physical limitations of space in a densely developed urban environment likely constrained further expansion, as new bikeways would require more complex and expensive measures, such as road reallocation or the creation of shared-use paths.
 
-Despite this slowdown in new installations, @fig-upgraded demonstrates a significant focus on upgrading existing infrastructure, particularly in later years.
-This shift from installation to upgrades could indicate a strategic pivot by the city toward improving the quality and safety of existing cycling infrastructure.
-As infrastructure ages and new safety standards emerge, upgrading older lanes becomes crucial to ensuring the network remains usable and safe for cyclists.
-The peak of upgrades in 2015, for example, might reflect the city's response to increased cycling activity, demands from advocacy groups, or the implementation of updated cycling policies that prioritize safer and more modern designs, such as protected bike lanes.
+Despite this slowdown in new installations, @fig-upgraded demonstrates a substantial focus on upgrading existing infrastructure, particularly in recent years.
+This shift from new installations to upgrades may indicate a strategic decision by the city to enhance the quality and safety of the current cycling infrastructure.
+As the infrastructure ages and new safety standards are introduced, upgrading older lanes is essential to ensuring the network remains functional and safe for cyclists.
+The peak of upgrades in 2015, for instance, may reflect the city's response to increased cycling activity, advocacy group demands, or the adoption of updated cycling policies that prioritize safer designs, such as protected bike lanes.
 
-This pattern of prioritizing upgrades over new installations also suggests that Toronto's initial expansion may have left some areas with suboptimal or unsafe infrastructure, particularly in the form of shared roadways or painted bike lanes that offer minimal protection to cyclists.
-As urban transportation planning evolves, the focus has shifted toward creating more separated and protected lanes, which are essential for encouraging cycling among less experienced riders, families, or those concerned about safety.
-This ongoing commitment to improving the infrastructure—rather than merely expanding it—reflects a maturation in Toronto’s approach to active transportation, as planners recognize that quality and safety are just as important as quantity when promoting cycling as a viable mode of transport.
+The shift toward prioritizing upgrades over new installations also suggests that the initial expansion left certain areas with inadequate or unsafe infrastructure, particularly in the form of shared roadways or painted bike lanes offering minimal cyclist protection.
+As urban transportation planning evolves, the emphasis has shifted toward constructing more segregated and protected lanes, which are vital for encouraging cycling among less experienced riders, families, or individuals concerned about safety.
+Considering new plans by the goverment [@citeCyclingNetworkPlan], this ongoing focus on improving infrastructure—rather than simply expanding it—reflects a more advanced approach to active transportation, recognizing that safety and quality are equally important to increasing cycling participation.
 
-Finally, another crucial aspect linked to the decline in installations could be budgetary constraints or competing infrastructure projects.
-As urban budgets are stretched across various priorities—public transit expansion, road maintenance, and green space development—the pace of installing new cycling lanes might slow.
-However, the investment in upgrades suggests that cycling infrastructure remains a priority, even if the focus has shifted from expansion to refinement.
-Continuing to monitor and analyze these trends will be essential for understanding how Toronto can balance its need for both new installations and the improvement of existing cycling routes to meet its long-term sustainability and mobility objectives.
+In conclusion, the analysis highlights the progression of Toronto’s cycling infrastructure from an initial phase of rapid expansion to a more recent emphasis on upgrading and enhancing the safety and quality of the network.
+This evolution reflects the complexities of balancing growth with maintenance, the need to adapt to changing urban environments, and the influence of international best practices and local advocacy in shaping the future of cycling in Toronto.
 
-In conclusion, the first discussion point highlights the evolution of Toronto’s cycling infrastructure from an initial phase of rapid expansion to a more recent focus on upgrading and enhancing the safety and quality of the network.
-This shift reflects the challenges of balancing growth with maintenance, the need to adapt to changing urban landscapes, and the influence of global best practices and local advocacy in shaping the city’s cycling future.
+## Comfort Level and Cyclist Safety
 
-## Second discussion point
+The classification of comfort levels in Toronto's cycling infrastructure offers significant insights into the city's capacity to ensure cyclist safety and encourage widespread usage.
+According to the Bikeway Comfort Level Distribution, @fig-comfort-one-1, a large portion of the network consists of lanes categorized as either High or Low Comfort, with Moderate Comfort lanes representing a smaller percentage.
+This distribution indicates that, although Toronto has expanded its cycling network, the infrastructure may still lack the protection required to encourage consistent cycling across a broader demographic.
+Ideally, Low Comfort lanes should form the smallest proportion, ensuring safer and more accessible routes for cyclists of all experience levels.
 
-The comfort level classification of Toronto's cycling infrastructure offers valuable insights into the quality of the city's bike network, particularly with respect to cyclist safety and usability.
-As shown in @fig-comfort-one, the majority of bikeways fall into the Low or High Comfort categories, with Moderate Comfort lanes—typically those offering relatively protected infrastructure—making up a smaller proportion of the overall network.
-This distribution suggests that, Toronto has successfully expanded the sheer number of bikeways, the majority could offer the protection and comfort necessary to encourage broader and more consistent usage.
-However, it is better to see Low Comfort level lanes taking the smallest portion among all.
+Comfort level classification is crucial for determining which segments of the population feel confident using the cycling network.
+Low Comfort lanes, including painted bike lanes and shared roadways, provide minimal physical separation from vehicular traffic.
+This limited protection often discourages risk-averse cyclists, particularly in areas with high traffic volumes.
+While such lanes may be acceptable in quieter residential areas, they are often avoided in busy urban environments.
+Unfortunately, as @fig-comfort-one-2 shows, 77.31% of these Low Comfort lanes remain unupgraded, revealing a substantial gap in the network's ability to offer safe and comfortable conditions for all cyclists.
 
-The classification of cycling lanes into different comfort levels is important because it directly affects who feels confident enough to cycle.
-Low Comfort lanes, often painted bike lanes or shared roadways, offer minimal physical separation from motor vehicles, making them less appealing to risk-averse cyclists.
-The lack of protection increases the perceived danger, discouraging people from cycling regularly, particularly in areas with high traffic volumes.
-For example, a cyclist may feel comfortable using a Low Comfort lane in a quiet residential neighborhood but avoid it entirely on a busy downtown street where traffic and intersections pose greater hazards.
+Moderate Comfort lanes, which provide a higher degree of separation but still lack full physical barriers, also face similar issues.
+As indicated in @fig-comfort-two-1, 76.62% of Moderate Comfort lanes have not been upgraded.
+This shortfall presents a significant challenge in attracting cyclists to busier, high-traffic areas where perceived risk is still considerable.
+While Moderate Comfort lanes are an improvement over Low Comfort options, they may not sufficiently address the safety concerns of groups particularly sensitive to danger, such as women, children, and older adults.
 
-Moderate Comfort lanes, which typically include designated bike lanes, offer more safety but still lack the physical separation provided by protected or High Comfort infrastructure.
-These lanes are a step in the right direction but may not sufficiently address concerns about safety, particularly in areas with fast-moving or congested vehicle traffic.
-Moderate Comfort lanes are likely to attract more cyclists than Low Comfort lanes, but they may not be enough to significantly increase ridership across all demographic groups.
-For instance, studies have shown that women, children, and older adults are particularly sensitive to perceived danger in cycling, and without sufficient protection, these groups are less likely to use the infrastructure regularly.
+In conclusion, while Toronto has made progress in expanding its cycling network, the high proportion of unupgraded lanes underscores the need for further improvements.
+The city should prioritize upgrading its Low and Moderate Comfort lanes while continuing to expand its High Comfort infrastructure.
+This shift will not only improve safety but also enhance equity in urban transportation, making cycling a more viable and appealing option for a diverse range of residents.
 
-High Comfort lanes, on the other hand, provide the greatest potential for encouraging cycling across a broad range of demographics.
-These lanes, often in the form of protected bike lanes or multi-use paths, offer a physically separated space for cyclists, reducing the likelihood of accidents involving motor vehicles.
-Research from cities that have invested heavily in High Comfort infrastructure, such as Amsterdam or Copenhagen, indicates that the availability of such protected lanes is a key factor in increasing overall cycling rates and making cycling a more viable and appealing transportation option for all.
-In Toronto, however, as the analysis shows, only a small proportion of the network is classified as High Comfort.
-This limits the extent to which the city can fully leverage its cycling infrastructure to promote widespread, sustained usage.
+## Geospatial Distribution and Expansion Challenges
 
-The low proportion of High Comfort lanes can be partly attributed to the early phases of Toronto's cycling infrastructure development, during which the focus was likely on quickly installing a large number of lanes to meet political and public demands.
-At that time, lower-cost solutions like painted bike lanes and sharrows (shared roadways) were easier and faster to implement than more expensive, space-consuming, and structurally complex protected lanes.
-However, these lanes did not necessarily provide the level of safety needed to support a significant increase in cycling activity, particularly among more cautious cyclists.
-The subsequent shift towards upgrading infrastructure, as discussed earlier, is likely an acknowledgment of the limitations of the initial approach, where quantity may have been prioritized over quality.
+The geospatial analysis of Toronto's cycling infrastructure, as shown in @fig-map, reveals both strengths and gaps in the city's bike network.
+One key observation is the dense concentration of bikeways in the downtown core.
+This concentration aligns with expectations, given that downtown Toronto is the city's most populated and congested area, with high demand for alternative transportation options.
+The extensive network of bikeways in this region reflects a targeted effort to make cycling a viable and accessible mode of transportation for residents, commuters, and visitors.
 
-Moreover, the analysis reveals that many of Toronto's Moderate and Low Comfort lanes have yet to be upgraded, as shown in @fig-comfort-one-2 and @fig-comfort-two.
-For both Low and Moderate Comfort lanes, over 75% have not been upgraded, which implies that much of the city's existing cycling network still operates under outdated or less protective standards.
-This points to a significant gap between the current state of Toronto's cycling infrastructure and the standards that are increasingly being recognized as necessary to make cycling a mainstream mode of transportation in urban environments.
-The failure to upgrade a larger portion of these lanes may stem from budgetary constraints, political will, or challenges in reallocating road space in already congested or car-dominated areas.
+However, @fig-map also reveals a significant disparity in infrastructure coverage as one moves away from the city center toward suburban and peripheral areas.
+In these regions, cycling infrastructure is sparse, underscoring unequal access to safe and efficient bike routes.
+This limited infrastructure in suburban areas reduces the overall utility of the city's cycling network, as it disproportionately benefits those residing in the urban core.
+Suburban residents, often facing longer commutes and a greater dependence on automobiles, would benefit from an expanded bike lane network, providing a safer and more convenient alternative to driving.
 
-One of the most critical implications of these findings is that Toronto’s cycling infrastructure might not yet be sufficiently robust to meet the city’s broader goals of reducing car dependency, lowering emissions, and promoting healthier, more active lifestyles.
-To truly shift a significant portion of the population towards cycling, the city will need to invest more heavily in upgrading its infrastructure, particularly in transitioning more of its Low and Moderate Comfort lanes into High Comfort lanes.
-This is especially important in areas where traffic volume and speed are higher, as these conditions pose greater risks to cyclists and deter people from choosing bikes as a mode of transportation.
+Several factors contribute to this discrepancy in cycling infrastructure.
+Suburban areas typically have lower population densities and roadways designed primarily for car travel, which historically has deprioritized cycling in urban planning.
+Additionally, suburban cultures often rely heavily on automobiles, leading to lower public demand for bike lanes.
+As Toronto aims to promote sustainable transportation and reduce car dependency, prioritizing the development of safe cycling routes in suburban areas will be essential.
+Expanding infrastructure in these regions can significantly contribute to the city’s broader transportation and environmental goals.
 
-Another potential reason for the focus on Moderate and Low Comfort lanes is the spatial distribution of the city’s bikeways.
-As mentioned earlier when showing the result of @fog-map, much of Toronto’s cycling network is concentrated in the downtown core, where road space is at a premium.
-Installing protected lanes in these areas often requires taking road space away from cars, which can be politically and technically challenging.
-In contrast, suburban areas, where road space is more abundant, could offer better opportunities for the city to build High Comfort infrastructure.
-Expanding the reach of the cycling network into these suburban areas, while upgrading the comfort levels of existing downtown lanes, could have a transformative effect on overall cycling rates and help make cycling a viable option for longer commutes.
+@fig-map also emphasizes the importance of recreational areas, particularly those near the waterfront, in shaping the cycling network.
+These areas often feature multi-use trails with High Comfort lanes, offering cyclists scenic and safe routes away from motor traffic.
+These trails serve not only as commuting routes but also as recreational spaces, promoting exercise and relaxation.
 
-In conclusion, while Toronto has made considerable progress in expanding its cycling infrastructure, the comfort level analysis reveals that much of the network remains suboptimal for ensuring safety and encouraging more widespread use.
-To achieve its sustainability and transportation goals, the city must focus on upgrading its existing Low and Moderate Comfort lanes and prioritize the construction of High Comfort infrastructure.
-This shift will be crucial not only for improving safety but also for making cycling an attractive option for a more diverse range of residents, thereby promoting equity and inclusivity in urban transportation.
+Finally, @fig-map points to opportunities for future expansion.
+Areas with few or no cycling routes should be considered for infrastructure development, especially where there is sufficient road space or public demand for safer cycling options.
+As Toronto continues to grow, addressing these gaps in the cycling network will be critical to ensuring that all residents, regardless of location, have access to safe, comfortable, and convenient cycling infrastructure.
+Expanding the network equitably across the city will support the development of a more inclusive and sustainable transportation system.
 
-## Third discussion point
-
-The geospatial analysis of Toronto's cycling infrastructure, as visualized in @fig-map, reveals clear spatial patterns that highlight both the strengths and weaknesses of the city’s current bike network.
-One of the key findings is the dense concentration of bikeways in the downtown core, where cycling infrastructure is most integrated into the existing urban fabric.
-This concentration aligns with expectations, as downtown Toronto is the city’s most densely populated and heavily trafficked area, where the demand for alternative transportation modes is highest.
-The high density of bikeways in the city center reflects a concerted effort to provide accessible cycling options in areas where they are most likely to be used, particularly by commuters and urban residents.
-
-However, the map also exposes a noticeable gap in cycling infrastructure as one moves further from the city center.
-The relative sparsity of bike lanes in suburban and peripheral areas suggests that while downtown residents enjoy a well-connected cycling network, those living in the suburbs have fewer options for safe, comfortable bike routes.
-This discrepancy may limit the overall effectiveness of Toronto’s cycling infrastructure, as it primarily serves those in the urban core while potentially neglecting a significant portion of the population.
-Suburban residents, who often have longer commutes and fewer public transit options, could benefit greatly from expanded cycling infrastructure that offers a viable alternative to car travel.
-
-The lack of infrastructure in suburban areas could be attributed to a variety of factors.
-Suburbs typically have lower population densities and wider roads, making cycling infrastructure less of a priority in urban planning compared to the densely populated downtown core.
-Additionally, suburban regions tend to have more car-dependent cultures, where cycling is not as ingrained as a common mode of transportation.
-The existing infrastructure may therefore reflect historical priorities that favored cars over active transportation.
-However, as cities like Toronto strive to reduce traffic congestion and meet climate goals, expanding cycling infrastructure into these suburban areas could be a crucial step toward encouraging more sustainable transportation choices.
-
-Furthermore, the map shows the importance of major recreational areas, such as parks and waterfronts, in Toronto’s cycling network.
-Multi-use trails, which provide High Comfort lanes, are particularly prevalent in these areas, offering scenic and safe routes for cyclists.
-While these trails are beneficial for recreational cycling, they may not be fully integrated into the city’s transportation network, limiting their utility for commuters.
-Connecting these trails to the downtown core and other major employment hubs could improve their role in promoting cycling as a legitimate form of daily transportation rather than just a leisure activity.
-
-Lastly, the map highlights potential future opportunities for infrastructure expansion.
-Areas with sparse cycling routes could be prioritized for new installations, especially where there is existing road capacity or public demand for safer cycling options.
-Identifying and addressing these gaps will be crucial as Toronto continues to grow and seeks to make its transportation system more equitable and sustainable.
-The current infrastructure is heavily concentrated in downtown areas, but future developments could focus on creating a more geographically balanced network, ensuring that all residents, regardless of location, have access to safe and convenient cycling routes.
-
-In summary, the geospatial distribution of Toronto’s cycling infrastructure reveals a well-connected downtown network but highlights the need for more investment in suburban areas and better integration of recreational trails into the broader transportation system.
-Expanding the reach of cycling infrastructure into underdeveloped areas will be key to making cycling a viable transportation option for a larger portion of the city’s population.
+In conclusion, while the spatial distribution of Toronto's cycling infrastructure demonstrates a well-connected downtown core, it also highlights the need for further investment in suburban regions.
+Expanding the cycling network in underserved areas will be crucial for promoting cycling as a practical and widely adopted mode of transportation throughout the city.
 

### 2024-09-24 scripts/02-data_cleaning.R
data cleaning updated

@@ -135,6 +135,7 @@ data <- read_csv("data/raw_data/raw_data.csv")
 lane_type_data <- data %>%
   select(INFRA_HIGHORDER, INFRA_LOWORDER, INSTALLED, UPGRADED) %>%   # Select relevant columns including year data
   filter(!is.na(INFRA_HIGHORDER) & !is.na(INFRA_LOWORDER)) %>%       # Remove rows with NA values
+  filter((INSTALLED >= 1800 | is.na(INSTALLED)) & (UPGRADED >= 1800 | is.na(UPGRADED))) %>%  # Remove rows where INSTALLED or UPGRADED equals 1
   mutate(
     Comfort_Level = case_when(                                       # Combine both columns into one classification
       str_detect(INFRA_HIGHORDER, "Protected|Multi-Use") | str_detect(INFRA_LOWORDER, "Protected|Multi-Use") ~ "High Comfort",
@@ -149,10 +150,5 @@ write_csv(
   x = lane_type_data,
   file = "data/analysis_data/lane_type_data.csv"
 )
-
-
-
-
-
 
 

### 2024-09-24 scripts/03-test_data.R
test added

@@ -8,8 +8,131 @@
 # Any other information needed? [...UPDATE THIS...]
 
 
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
 
-#### Test data ####
+#### Workspace setup ####
+library(here)
+library(dplyr)
+library(testthat)
+library(lubridate)
+library(geojsonio)
+library(tidyverse)
+
+
+
+#### Test Raw Data ####
+# Test suite
+test_that("Test for missing values", {
+  data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+  missing_info <- sapply(data, function(x) sum(is.na(x)))
+  missing_cols <- names(missing_info[missing_info > 0])
+  expect_true(length(missing_cols) > 0, info = "Columns with missing values exist.")
+  print(missing_cols)
+})
+
+test_that("Test for valid data types", {
+  # Check if INSTALLED, UPGRADED, OBJECTID are numeric
+  data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+  numeric_check <- all(sapply(data[c('INSTALLED', 'UPGRADED', 'OBJECTID')], is.numeric))
+  expect_true(numeric_check, info = "INSTALLED, UPGRADED, OBJECTID should be numeric.")
+  
+  # Check if LAST_EDIT_DATE has a valid date format (using ymd_hms for ISO 8601)
+  date_check <- all(!is.na(ymd_hms(data$LAST_EDIT_DATE)))
+  expect_true(date_check, info = "LAST_EDIT_DATE should be in valid date format.")
+})
+
+test_that("Test for valid ranges of years", {
+  data <- read_csv(here("data", "raw_data", "raw_data.csv"))
+  current_year <- year(Sys.Date())
+  
+  # Check INSTALLED range
+  installed_valid <- all(data$INSTALLED >= 1900 & data$INSTALLED <= current_year, na.rm = TRUE)
+  expect_true(installed_valid, info = "INSTALLED year should be between 1900 and the current year or NA.")
+  
+  # Check UPGRADED range, allow NA values
+  upgraded_valid <- all(is.na(data$UPGRADED) | (data$UPGRADED >= 1900 & data$UPGRADED <= current_year | data$UPGRADED == 1)) #there is a invalid UPGRADED data 1
+  expect_true(upgraded_valid, info = "UPGRADED year should be between 1900 and the current year or NA.")
+})
+
+test_that("Test for uniqueness of X_id and SEGMENT_ID", {
[…]

### 2024-09-24 paper/paper.qmd
paper first draft-minor update

@@
 @fig-comfort-one provides a breakdown of Toronto’s bikeway network by comfort level and upgrade status.
-@fig-comfort-one-1 shows the distribution of all bikeways based on comfort level, with the majority of bikeways classified as either Low Comfort (41.6%) or Moderate Comfort (41.1%), while High Comfort lanes constitute a smaller portion (17.3%).
+@fig-comfort-one-1 shows the distribution of all bikeways based on comfort level, with the majority of bikeways classified as either Low Comfort (40.88%) or Moderate Comfort (41.75%), while High Comfort lanes constitute a smaller portion (17.36%).
 This indicates that while Toronto has invested in a significant number of bike lanes, most of them fall within lower comfort categories, suggesting that further improvements, particularly in upgrading existing lanes, are needed to promote safer and more comfortable cycling experiences.
@@
 @fig-comfort-one-2, @fig-comfort-two-1, and @fig-comfort-two-2 offer a deeper look into the upgrade status for each comfort level.
-For Low Comfort lanes, only 22.69% have been upgraded, while 77.31% remain in their original state.
+For Low Comfort lanes, only 22.26% have been upgraded, while 77.74% remain in their original state.
 Similarly, for Moderate Comfort bikeways, 23.38% have been upgraded, leaving 76.62% untouched.
@@
 While such lanes may be acceptable in quieter residential areas, they are often avoided in busy urban environments.
-Unfortunately, as @fig-comfort-one-2 shows, 77.31% of these Low Comfort lanes remain unupgraded, revealing a substantial gap in the network's ability to offer safe and comfortable conditions for all cyclists.
+Unfortunately, as @fig-comfort-one-2 shows, 77.74% of these Low Comfort lanes remain unupgraded, revealing a substantial gap in the network's ability to offer safe and comfortable conditions for all cyclists.
 

### 2024-09-25 paper/paper.qmd
paper second draft

@@
 date-format: long
-abstract: "This paper explores the development of Toronto's cycling infrastructure.
-Through a comprehensive analysis of installation dates, upgrades, and comfort levels of bike lanes, this study highlights the trends in cycling infrastructure expansion from 2001 to 2024.
-A significant focus is placed on the comfort levels of different types of lanes, offering insights into how the infrastructure meets the needs of cyclists.
-The spatial distribution of bike lanes is also analyzed, revealing discrepancies in access between downtown and suburban areas.
-The findings emphasize the need for continued investment in both the quality and expansion of cycling infrastructure to promote sustainable transportation in Toronto."
+abstract: "This study examines the evolution of Toronto's cycling infrastructure between 2001 and 2024, focusing on installation trends, upgrades, and comfort levels of bike lanes.
+The analysis reveals that while the cycling network has grown, many lanes remain low in comfort, with significant disparities in access between downtown and suburban areas.
+These findings highlight the need for continued investment in upgrading existing lanes and expanding the network to promote safer and more equitable cycling options across the city.
+Such improvements are essential for supporting sustainable transportation and reducing car dependency in Toronto."
 format: pdf
@@
 
-# Load packages
+# Package List
 package_list <- c("sf", "readr", "ggplot2", "osmdata", "tidyverse", "jsonlite", "dplyr", "here", "knitr", "kableExtra")
 
-# Function to check and install missing packages
+# Check and install missing packages
 install_and_load <- function(package_list) {
@@
 install_and_load(package_list)
+
 ```
@@
 
-Urban cycling plays a critical role in advancing sustainable transportation strategies, especially in rapidly growing cities like Toronto.
-With the city facing increasing congestion, promoting active transportation modes such as cycling is essential for reducing greenhouse gas emissions, improving public health, and enhancing urban mobility.
-Since the early 2000s, Toronto has made significant strides in expanding its cycling infrastructure, creating a network of bike lanes, multi-use paths, and shared roadways to support cyclists citywide.
+The development of urban cycling infrastructure is critical for promoting sustainable transportation, reducing greenhouse gas emissions, and improving public health.
+In rapidly growing cities like Toronto, the integration of cycling lanes into the urban transportation system provides an alternative to car dependency, addressing concerns related to traffic congestion and environmental sustainability.
+However, despite Toronto’s growing network of bike lanes, there remains significant variation in the quality and accessibility of these lanes, particularly between downtown and suburban areas.
+For most Toronto residents, they may see some bicycles riding on the road everyday, but there is a high probability that there is no clear understanding of Toronto's bike lanes construction.
+This paper is meant to fulfill this knowledge gap and give people an overall understanding on Toronto's cycling network development and why it is important.
 
-This paper analyzes the evolution of Toronto's cycling infrastructure over the past two decades, focusing on the installation and upgrade trends of bikeways and their classification by comfort level.
-Using the 'Cycling Network' dataset from Open Data Toronto, the study examines both the spatial distribution of bike lanes and the varying levels of cycling comfort across different regions.
-The analysis highlights discrepancies between downtown and suburban areas, showcasing progress while identifying regions that remain underserved.
+This paper focuses on the evolution of Toronto’s cycling infrastructure from 2001 to 2024, providing a geospatial and temporal analysis of bikeway installations, upgrades, and comfort levels across the city.
+Using the publicly available 'Cycling Network' dataset [@citeCyclingNetwork] from Open Data Toronto [@citeODT], this study investigates the distribution of bike lanes, the different lanes' comfort levels with its classifications, and the trends in cycling infrastructure development over the past two decades.
+It highlights the differences in the network's coverage between the densely populated downtown core and the more sparsely served suburban regions.
 
-The study's findings underscore the need for continued investment in both the expansion and quality of Toronto's cycling infrastructure to encourage more widespread adoption of cycling as a viable mode of transportation.
-By addressing both the geographic imbalances and comfort levels of bike lanes, the city can better meet its sustainability goals.
+A critical gap in the literature is the lack of detailed, data-driven studies assessing the quality of Toronto’s cycling infrastructure.
+While the city has made strides in expanding its network, many lanes remain of low comfort, which may discourage public choosing cycling as a daily mode of transportation.
+This paper aims to solve this problem by analyzing both the growth of the network and its current state in terms of usability and safety.
 
-The structure of this paper is as follows: the data sources and methodologies are presented in @sec-data, followed by an analysis of the results in @sec-res, and a discussion of the findings in @sec-dis.
+To address this, the study evaluates the installation and upgrade trends of bike lanes, classifies the comfort levels of these lanes, and identifies areas where further investment is necessary.
+The results indicate that while Toronto’s cycling network has expanded, many lanes offer minimal physical protection, with significant underinvestment in suburban areas.
+These findings underscore the need for continuous improvements to both the safety and accessibility of the network, which are essential for supporting Toronto’s broader goals of sustainable urban mobility.
+
+The importance of this study lies in its potential to inform urban planners and policymakers about the current shortcomings of Toronto’s cycling infrastructure.
+By emphasizing the need for targeted investments, particularly in underserved regions, the study contributes to ongoing efforts to make cycling a more viable and equitable transportation option across the city.
+
+The structure of this paper is organized as follows: @sec-data presents the data sources and methodologies, followed by an overview of the results in @sec-res.
+The findings are then discussed in @sec-dis, and additional detailed information is provided in the appendix at @sec-app.
 
 # Data {#sec-data}
+
+This project is structured based on Rohan's principles [@citeRohan].
+Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR].
+Also with support of additional packages in R: `sf` [@citeSf], `readr` [@citeReadr], `ggplot2` [@citeGgplot2], `osmdata` [@citeOsmdata], `tidyverse` [@citeTidyverse], `jsonlite` [@citeJsonlite], `dplyr` [@citeDplyr], `here` [@citeHere], `knitr` [@citeKnitr], `kableExtra` [@citeKableExtra].
 
@@
 This dataset is publicly available and is continually updated to reflect Toronto’s ongoing efforts to develop and maintain its cycling infrastructure.
-It contains geospatial and temporal information about various cycling-related features, including the types of bikeways (dedicated lanes, shared roads, multi-use trails), installation years, and any subsequent upgrades.
-By offering a detailed history of Toronto's cycling network development, this dataset plays a vital role in informing both urban planning policies and research into sustainable transportation options.
+It contains geospatial and temporal information about various cycling-related features, including the types of bikeways, installation years, etc.
+Such a detailed history of Toronto's cycling network development shows a good fit for this study.
 
 While the Cycling Network dataset was the primary source of data, other relevant datasets were explored, such as the Road Network File [@citeStatCan] from Statistics Canada, which provides detailed geospatial data on all roadways in Toronto.
-Another such alternative was the Major City-Wide Cycling Routes dataset [@citeCyclingRoutes], also available on Open Data Toronto [@citeODT].
+Another alternative was the Major City-Wide Cycling Routes dataset [@citeCyclingRoutes], also available on Open Data Toronto [@citeODT].
 However, this dataset focuses specifically on the large-scale cycling routes connecting different parts of the city, emphasizing connectivity at a macro level.
-While these two dataset are useful for analyzing high-level transportation patterns, it lacks the detailed, segment-level data—such as comfort levels, installation dates, and upgrade history—that are crucial for a deep analysis of cycling infrastructure.
+While these two dataset are useful for analyzing high-level transportation patterns, it lacks the detailed data such as comfort levels, installation dates, and upgrade history, that are crucial for a deep analysis of cycling infrastructure.
 Since this paper focuses on both the growth and quality of cycling infrastructure, the Cycling Network dataset, with its more comprehensive coverage of individual bikeway segments, was a better fit for the goals of this study.
-
-Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR]. Also with support of additional packages which will be mentioned later in the @sec-data-method.
 
@@
 
-The 'Cycling Network' dataset from Open Data Toronto tracks various aspects of Toronto's cycling infrastructure, such as installation dates, upgrades, street names, and the type of infrastructure (e.g., sharrows, multi-use trails).
-Each entry represents a segment of the cycling network, where real-world phenomena, such as the construction or upgrading of cycling paths, are documented.
-For example, when a cycling path is installed or upgraded, the responsible authorities collect information such as the installation year (e.g., 2001) and any subsequent upgrades (e.g., 2021).
-This data is then digitized into structured entries in the dataset.
+The 'Cycling Network' dataset from Open Data Toronto tracks various aspects of Toronto's cycling infrastructure, such as installation dates, upgrades, street names, and the type of the lane.
+Each entry represents a segment of the cycling network, such as the construction or upgrading of cycling paths, are documented.
+For example, when a cycling path is installed or upgraded, the responsible authorities collect information such as the installation year and any year of subsequent upgrades.
+This data is then inputed into structured entries in the dataset.
 
-To go from a physical event (e.g., the installation of a bike lane) to a dataset entry, detailed records are maintained by the city's transportation department.
-These records are geospatially coded (column geometry in this dataset), allowing each cycling path to be mapped precisely in relation to other city infrastructure.
+To go from a physical event, such as installing a new lane, to a dataset entry, detailed records are maintained by the city's transportation department.
+These records include precise latitude and longitude of the lanes, allowing each cycling path to be mapped precisely in relation to other city infrastructure.
 Thus, the phenomena of constructing or updating a bike lane becomes an entry with attributes such as the street name, road class, and installation history.
@@
 
-As mentioend above, the dataset contains several key variables that were used in this analysis:
+The various variables mentioned above serves as key variables that were used in analysis for this paper:
 
@@
 #### Comfort Level
+
 Each bikeway is assigned a comfort level—high, moderate, or low—based according to the recorded Bikeway Type.
@@
 ## Method {#sec-data-method}
-
-Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR].
-Also with support of additional packages in R: `sf` [@citeSf], `readr` [@citeReadr], `ggplot2` [@citeGgplot2], `osmdata` [@citeOsmdata], `tidyverse` [@citeTidyverse], `jsonlite` [@citeJsonlite], `dplyr` [@citeDplyr], `here` [@citeHere], `knitr` [@citeKnitr], `kableExtra` [@citeKableExtra].
 
@@
 
-In the first data cleaning process, the focus was on extracting and cleaning geographical coordinates from the 'geometry' column, which contains location data in JSON-like format, @tbl-raw-coordinates displays how the data was like.
-A custom function was created to parse the coordinates, filter out invalid entries, and extract the longitude and latitude values.
-The extracted coordinates were combined into a new dataset shon in @tbl-clean-coordinates, with unique IDs to match the original entries.
+In the first data cleaning process, the focus was on extracting and cleaning geographical coordinates from the 'geometry' column, which all data are in JSON-like format, @tbl-raw-coordinates displays how the raw data was like.
+A function was created to parse the coordinates, filter out invalid entries, and extract the longitude and latitude values.
+The extracted coordinates were combined into a new dataset shown in @tbl-clean-coordinates, with unique IDs to match the original cycling lanes they from.
 Any rows with missing coordinate data were removed, and the cleaned data was saved into a separate file for further analysis.
@@
 Data from the 'INFRA_HIGHORDER' and 'INFRA_LOWORDER' columns was cleaned, and missing values were removed.
-The lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, depending on their Bikeway Type data, detail of this classification can be found in @sec-app-comfort.
+By using the Bikeway Type data, the lanes were classified into three comfort levels: High Comfort, Moderate Comfort, and Low Comfort, detail of this classification can be found in @sec-app-comfort.
 The classification, along with year data for installation and upgrades, was saved for analysis on the types of cycling lanes and their comfort levels.
@@
 A significant spike is observed in 2001, with over 400 bikeways installed in that year alone.
-This likely reflects a major investment in cycling infrastructure at the time.
+This likely reflects the starting time of this dataset, which all lanes built before 2001 was recorded as 2001.
 Following 2001, the number of installations drops sharply and stabilizes at much lower levels in subsequent years, with occasional small peaks around 2007, 2011, and 2021.
@@
 
-@fig-upgraded illustrates the number of bikeways upgraded in Toronto from 2001 to 2023, with panel (a) showing the raw counts of upgrades and panel (b) visualizing the trend.
+@fig-upgraded illustrates the number of bikeways upgraded in Toronto from 2001 to 2024, with panel (a) showing the raw counts of upgrades and panel (b) visualizing the trend.
 @fig-upgraded-1 reveals that the most significant number of upgrades occurred in 2015, where nearly 100 bikeways were updated, marking a major peak.
-Outside of this spike, the number of annual upgrades remains relatively modest, typically fluctuating between 10 and 50 bikeways.
+Outside of this spike, the number of annual upgrades remains relatively modest, typically fluctuating between 10 and 50 bikeways each year.
 Notable periods of increased upgrades are visible around 2012 and 2020, possibly reflecting concentrated efforts to modernize or expand existing cycling infrastructure in those years.
@@
 In @fig-upgraded-2, the overall trend is slightly upward, as indicated by the red trend line.
-Despite the notable peaks and valleys, the trend line suggests that, on average, the number of bikeway upgrades has increased over the years.
+Despite the notable peaks and drops, the trend line suggests that, on average, the number of bikeway upgrades has increased over the years.
 The spike in 2015 significantly affects the trend, but even excluding that, the gradual rise suggests that more consistent efforts have been made to maintain and improve Toronto’s cycling network.
 This trend aligns with broader urban planning strategies focused on making cities more accessible and environmentally friendly by supporting active transportation.
-The dips in recent years, particularly after 2020, might be influenced by resource reallocation or external factors, such as the global pandemic, which could have delayed infrastructure projects.
+The dips in recent years, particularly after 2020, might be influenced by resource reallocation or external factors, such as the global pandemic, which could have delayed some plans.
 
@@
 Similarly, for Moderate Comfort bikeways, 23.38% have been upgraded, leaving 76.62% untouched.
-Interestingly, High Comfort lanes have the lowest proportion of upgrades, with only 18.76% improved, despite representing the most comfortable and potentially most desirable biking infrastructure.
-These trends highlight a potential gap in Toronto's focus on maintaining and upgrading its most comfortable and accessible cycling lanes, suggesting a need for more targeted investments in high-comfort bikeways to encourage broader use of cycling for transportation.
+Interestingly, High Comfort lanes have the lowest proportion of upgrades, with only 18.76% improved.
+This result shows not much cycling lanes achieved high comfort level by upgrading, most of them are initially designed and built to be high comfort lanes.
+These trends highlight a potential gap in Toronto's focus on maintaining and upgrading its most comfortable and accessible cycling lanes, suggesting a need for more targeted investments in high-comfort bikeways to encourage cycling in the city.
 
@@
 @fig-map provides a visual representation of the bike lane infrastructure across Toronto.
-The lines on the map correspond to various cycling paths, including dedicated bike lanes, multi-use paths, and shared roadways, which span different areas of the city.
+The lines on the map correspond to various cycling paths, which span different areas of the city.
 The geographic coordinates, ranging from around 43.60°N to 43.85°N in latitude and -79.6°W to -79.3°W in longitude, encompass most of Toronto’s urban region.
@@
 
-A key observation from the map is the dense concentration of bike lanes in the downtown core and along major roadways, indicating a well-established network in central Toronto.
+A key observation from the map is the dense concentration of bike lanes in the downtown areas and along major roadways, indicating a well-established network in central Toronto.
 The downtown area’s grid-like structure is particularly evident, showcasing a high density of lanes that facilitate easy cycling access within the city's busiest areas.
 As one moves further away from the city center, the map shows fewer bike lanes, suggesting that suburban regions may have less developed cycling infrastructure.
-However, there are notable multi-use paths extending toward the outskirts, particularly in parks and recreational areas, providing important cycling connections for those commuting or using bikes for recreational purposes.
 
@@
 The data reveals a phased development of Toronto's cycling infrastructure, with a notable increase in installations during the early 2000s, as shown in @fig-installed.
-This rise is likely due to the dataset commencing in 2001, thereby consolidating all installations prior to this year.
-Additionally, the onset of formal record-keeping and the significant number of installations during this period may reflect political momentum and public demand for environmentally sustainable transportation alternatives, influenced by growing concerns over climate change and urban congestion.
+As mentioned before, this rise is likely due to the dataset starting in 2001, thereby recorded as 2001 for all installations prior to this year.
+However, if there was a large numebr of lanes installed in 2001, then it may reflect political momentum and public demand for environmentally sustainable transportation alternatives, influenced by growing concerns over climate change and urban congestion.
 
-Ignoring the data from the first year, the annual number of new bikeway installations still showing a trend of slight decreasing.
+Ignoring the data from the first year, the annual number of new bikeway installations will still show a trend of slight decreasing.
 Several factors may account for this decline.
-In the early stages, the city's cycling plan likely targeted areas where infrastructure could be installed with relative ease, particularly in downtown Toronto, where road networks and urban density favored such investments.
+In the early stages, the city's cycling plan likely targeted areas where infrastructure could be installed with relative low difficulties, particularly in downtown Toronto, where road networks and urban density favored such investments.
 As the network extended into more complex regions, such as suburban areas with fewer existing roadways, the rate of installation slowed.
@@
 As the infrastructure ages and new safety standards are introduced, upgrading older lanes is essential to ensuring the network remains functional and safe for cyclists.
-The peak of upgrades in 2015, for instance, may reflect the city's response to increased cycling activity, advocacy group demands, or the adoption of updated cycling policies that prioritize safer designs, such as protected bike lanes.
+Also, some study have shown the possibilities of improving traffic issues by upgrading the existing cycling lanes [@citeCyclingImprove], this study also mentioned increasing the accessibility can highly impact people's will to choose cycling to commute.
 
 The shift toward prioritizing upgrades over new installations also suggests that the initial expansion left certain areas with inadequate or unsafe infrastructure, particularly in the form of shared roadways or painted bike lanes offering minimal cyclist protection.
-As urban transportation planning evolves, the emphasis has shifted toward constructing more segregated and protected lanes, which are vital for encouraging cycling among less experienced riders, families, or individuals concerned about safety.
-Considering new plans by the goverment [@citeCyclingNetworkPlan], this ongoing focus on improving infrastructure—rather than simply expanding it—reflects a more advanced approach to active transportation, recognizing that safety and quality are equally important to increasing cycling participation.
-
-In conclusion, the analysis highlights the progression of Toronto’s cycling infrastructure from an initial phase of rapid expansion to a more recent emphasis on upgrading and enhancing the safety and quality of the network.
-This evolution reflects the complexities of balancing growth with maintenance, the need to adapt to changing urban environments, and the influence of international best practices and local advocacy in shaping the future of cycling in Toronto.
+As urban transportation planning evolves, the emphasis has shifted toward constructing more segregated and protected lanes, which are vital for encouraging cycling among less experienced riders or individuals concerned about safety.
+Considering new plans by the goverment [@citeCyclingNetworkPlan], this ongoing focus on improving infrastructure rather than simply expanding it, reflects a more advanced approach to active transportation, recognizing that safety and quality are equally important to increasing cycling participation.
 
 ## Comfort Level and Cyclist Safety
-
-The classification of comfort levels in Toronto's cycling infrastructure offers significant insights into the city's capacity to ensure cyclist safety and encourage widespread usage.
-According to the Bikeway Comfort Level Distribution, @fig-comfort-one-1, a large portion of the network consists of lanes categorized as either High or Low Comfort, with Moderate Comfort lanes representing a smaller percentage.
-This distribution indicates that, although Toronto has expanded its cycling network, the infrastructure may still lack the protection required to encourage consistent cycling across a broader demographic.
-Ideally, Low Comfort lanes should form the smallest proportion, ensuring safer and more accessible routes for cyclists of all experience levels.
 
@@
 
-In conclusion, while Toronto has made progress in expanding its cycling network, the high proportion of unupgraded lanes underscores the need for further improvements.
-The city should prioritize upgrading its Low and Moderate Comfort lanes while continuing to expand its High Comfort infrastructure.
-This shift will not only improve safety but also enhance equity in urban transportation, making cycling a more viable and appealing option for a diverse range of residents.
+Moreover, a study have shown the importance of buffer zones in cycling lanes [@citeLaneBuffer], it not only provide higher safety level, but also give people higer confidence to use these lanes when there are vehicles on the road.
+According to the Bikeway Comfort Level Distribution, @fig-comfort-one-1, a large portion of the network consists of lanes categorized as either High or Low Comfort, with Moderate Comfort lanes representing the samllest percentage.
+This distribution indicates that, although Toronto has expanded its cycling network, the infrastructure may still lack the high quality cycling lanes to encourage consistent cycling across mroe people.
+Ideally, Low Comfort lanes should form the smallest proportion, ensuring safer and more accessible routes for cyclists of all experience levels.
+In such cases, combining what we saw in the buffer zone study, there will be more usage of cycling lanes, thus help Toronto become a healthier and more environmental friendly city.
 
@@
 The geospatial analysis of Toronto's cycling infrastructure, as shown in @fig-map, reveals both strengths and gaps in the city's bike network.
-One key observation is the dense concentration of bikeways in the downtown core.
-This concentration aligns with expectations, given that downtown Toronto is the city's most populated and congested area, with high demand for alternative transportation options.
-The extensive network of bikeways in this region reflects a targeted effort to make cycling a viable and accessible mode of transportation for residents, commuters, and visitors.
+One key observation is the dense concentration of bikeways in the downtown Toronto.
+This concentration aligns with expectations, given that downtown areas are the city's most populated and crowded area, with high demand for alternative transportation options.
+The extensive network of bikeways in this region reflects a targeted effort to make cycling a viable and accessible mode of transportation for residents and visitors.
 
-However, @fig-map also reveals a significant disparity in infrastructure coverage as one moves away from the city center toward suburban and peripheral areas.
+However, @fig-map also reveals a significant disparity in infrastructure coverage as one moves away from the city center toward suburban areas.
 In these regions, cycling infrastructure is sparse, underscoring unequal access to safe and efficient bike routes.
-This limited infrastructure in suburban areas reduces the overall utility of the city's cycling network, as it disproportionately benefits those residing in the urban core.
-Suburban residents, often facing longer commutes and a greater dependence on automobiles, would benefit from an expanded bike lane network, providing a safer and more convenient alternative to driving.
-
+This limited infrastructure in suburban areas reduces the overall utility of the city's cycling network, as it cannot connect to all areas that you may want to go.
 Several factors contribute to this discrepancy in cycling infrastructure.
-Suburban areas typically have lower population densities and roadways designed primarily for car travel, which historically has deprioritized cycling in urban planning.
-Additionally, suburban cultures often rely heavily on automobiles, leading to lower public demand for bike lanes.
+Suburban areas typically have lower population densities and roadways designed primarily for car travel, which historically has lowered the priority of cycling infrastructure.
+However, expand cycling lanes to these areas would still benefit people since it provids an alternative choice other than driving.
+This can also help those who haven't get a driver's license to have more flexible ways of transportation.
 As Toronto aims to promote sustainable transportation and reduce car dependency, prioritizing the development of safe cycling routes in suburban areas will be essential.
-Expanding infrastructure in these regions can significantly contribute to the city’s broader transportation and environmental goals.
+Expanding infrastructure in these regions can significantly contribute to the city’s broader transportation and environmental goals [@citeEnvGoal].
 
 @fig-map also emphasizes the importance of recreational areas, particularly those near the waterfront, in shaping the cycling network.
-These areas often feature multi-use trails with High Comfort lanes, offering cyclists scenic and safe routes away from motor traffic.
+These areas often feature multi use trails with High Comfort lanes, offering cyclists scenic and safe routes away from motor traffic.
 These trails serve not only as commuting routes but also as recreational spaces, promoting exercise and relaxation.
@@
 As Toronto continues to grow, addressing these gaps in the cycling network will be critical to ensuring that all residents, regardless of location, have access to safe, comfortable, and convenient cycling infrastructure.
-Expanding the network equitably across the city will support the development of a more inclusive and sustainable transportation system.
-
-In conclusion, while the spatial distribution of Toronto's cycling infrastructure demonstrates a well-connected downtown core, it also highlights the need for further investment in suburban regions.
-Expanding the cycling network in underserved areas will be crucial for promoting cycling as a practical and widely adopted mode of transportation throughout the city.
+Lastly, it was mentioned as the first choice in the 'Low Carbon Transportation' article [@citeLowCarb] by City of Toronto, it further emphasized the importance of axpanding the network across the city to support the development of a more inclusive and sustainable transportation system.
 
@@
 
-One key limitation of this study is the reliance on the 'Cycling Network' dataset from OpenDataToronto, which, while comprehensive, might not capture informal or emerging cycling routes that are not officially recognized.
+One key limitation of this study is the reliance on the 'Cycling Network' dataset from Open Data Toronto, which, while comprehensive, might not capture informal or emerging cycling routes that are not officially recognized.
 Additionally, the dataset may not include newer infrastructure developments if updates were delayed in publication.
-Future research should integrate real-time data collection techniques or alternative datasets (e.g., crowdsourced information) to ensure that the analysis reflects the most current state of the network.
+Future research should integrate real-time data collection techniques or alternative datasets such as crowdsourced information to ensure that the analysis reflects the most current state of the network.
 
-Another weakness lies in the classification of comfort levels, which, while useful, is a simplification.
-Comfort in cycling can be influenced by numerous factors beyond infrastructure type, such as traffic density, road conditions, and seasonal weather.
-A more nuanced classification that incorporates these additional factors could provide a more accurate assessment of how comfortable Toronto's cycling network truly is for different types of cyclists.
+Another weakness lies in the classification of comfort levels, which is  useful but is an over simplification of this problem.
+Comfort in cycling can be influenced by numerous factors beyond infrastructure type, such as traffic density, road conditions, etc.
+A more comprehensive classification that incorporates these additional factors could provide a more accurate assessment of how good Toronto's cycling network truly is.
 

### 2024-09-26 scripts/00-simulate_data.R
scripts preamble section updated

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates a dataset representing Toronto's cycling network with route attributes such as length, surface type, and average daily riders.
+# Author: Ziheng Zhong
+# Date: 23 September 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: NA
+# Any other information needed: NA
 
 
 

### 2024-09-26 scripts/01-download_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from the Toronto Open Data cycling network package.
+# Author: Ziheng Zhong
+# Date: 23 September 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: NA
+# Any other information needed: NA
 
 
 

### 2024-09-26 scripts/02-data_cleaning.R

@@ -1,10 +1,11 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers
-# Author: Justin [...UPDATE THIS...]
-# Date: 21 September 2024
-# Contact: justin@example.com [...UPDATE THIS...]
+# Purpose: Cleans and processes cycling network data for analysis, including extracting coordinates, summarizing upgrades and installations, and classifying lane types by comfort level.
+# Author: Ziheng Zhong
+# Date: 23 September 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Dataset with geometry field containing valid JSON-like data
+# Pre-requisites: NA
+# Any other information needed: NA
 
 
 

### 2024-09-26 scripts/03-test_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests the integrity of raw and processed cycling network data, including checks for missing values, valid data types, and year ranges.
+# Author: Ziheng Zhong
+# Date: 23 September 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: NA
+# Any other information needed: NA
 
 
 

### 2024-09-26 scripts/04-charts.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Models and visualizes Toronto's cycling infrastructure data, including maps and charts on bikeway installations, upgrades, and comfort levels.
+# Author: Ziheng Zhong
+# Date: 23 September 2024
+# Contact: ziheng.zhong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: NA
+# Any other information needed: NA
 
 
 #### Workspace setup ####

### 2024-09-26 scripts/00-simulate_data.R
scripts processed by styler

@@ -21,17 +21,15 @@ set.seed(304)
 
 # Create a simulated dataset for Toronto Cycling Network
 toronto_cycling_network <- data.frame(
-  route_id = 1:100,  # Unique identifier for each cycling route
-  route_name = paste("Route", 1:100),  # Route name
-  length_km = round(rgamma(100, shape = 2, scale = 2), 2),  # Gamma distribution for route length (right-skewed)
-  surface_type = sample(c("Asphalt", "Concrete", "Gravel"), 100, replace = TRUE, prob = c(0.6, 0.3, 0.1)),  # Asphalt is more common
-  ward = sample(1:25, 100, replace = TRUE, prob = dpois(1:25, lambda = 12) + 0.01),  # Poisson distribution for ward, centered around 12
-  year_installed = sample(1990:2023, 100, replace = TRUE, prob = dnorm(1990:2023, mean = 2010, sd = 8)),  # More routes in recent years
-  maintenance_status = sample(c("Good", "Fair", "Poor"), 100, replace = TRUE, prob = c(0.5, 0.3, 0.2)),  # 50% chance of "Good" maintenance
-  avg_daily_riders = round(rlnorm(100, meanlog = log(300), sdlog = 0.4))  # Log-normal distribution for daily riders
+  route_id = 1:100, # Unique identifier for each cycling route
+  route_name = paste("Route", 1:100), # Route name
+  length_km = round(rgamma(100, shape = 2, scale = 2), 2), # Gamma distribution for route length (right-skewed)
+  surface_type = sample(c("Asphalt", "Concrete", "Gravel"), 100, replace = TRUE, prob = c(0.6, 0.3, 0.1)), # Asphalt is more common
+  ward = sample(1:25, 100, replace = TRUE, prob = dpois(1:25, lambda = 12) + 0.01), # Poisson distribution for ward, centered around 12
+  year_installed = sample(1990:2023, 100, replace = TRUE, prob = dnorm(1990:2023, mean = 2010, sd = 8)), # More routes in recent years
+  maintenance_status = sample(c("Good", "Fair", "Poor"), 100, replace = TRUE, prob = c(0.5, 0.3, 0.2)), # 50% chance of "Good" maintenance
+  avg_daily_riders = round(rlnorm(100, meanlog = log(300), sdlog = 0.4)) # Log-normal distribution for daily riders
 )
 
 # Check the dataset
 head(toronto_cycling_network)
-
-

### 2024-09-26 scripts/01-download_data.R

@@ -24,8 +24,8 @@ package <- show_package("cycling-network")
 resources <- list_package_resources("cycling-network")
 
 # choose .csv to download
-datastore_resources <- filter(resources, tolower(format) %in% c('csv'))
-raw_data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+datastore_resources <- filter(resources, tolower(format) %in% c("csv"))
+raw_data <- filter(datastore_resources, row_number() == 1) %>% get_resource()
 
 
 
@@ -34,5 +34,3 @@ write_csv(
   x = raw_data,
   file = "data/raw_data/raw_data.csv"
 )
-
-

### 2024-09-26 scripts/02-data_cleaning.R

@@ -28,22 +28,25 @@ extract_coordinates <- function(geo) {
   if (!is.na(geo)) {
     # Clean and prepare the JSON-like structure
     geo_clean <- gsub("'", "\"", geo)
-    
+
     # Convert to JSON and extract coordinates
-    json_data <- tryCatch({
-      fromJSON(geo_clean)
-    }, error = function(e) {
-      return(NULL)
-    })
-    
+    json_data <- tryCatch(
+      {
+        fromJSON(geo_clean)
+      },
+      error = function(e) {
+        return(NULL)
+      }
+    )
+
     if (!is.null(json_data) && !is.null(json_data$coordinates)) {
       # Flatten coordinates (assuming it's a nested list)
       coords <- unlist(json_data$coordinates, recursive = TRUE)
-      
+
       # Separate longitude (starting with -79) and latitude (starting with 43)
       longitude <- coords[grepl("^-79", coords)]
       latitude <- coords[grepl("^43", coords)]
-      
+
       # Ensure we have pairs
       if (length(longitude) == length(latitude)) {
         coords_matrix <- cbind(longitude, latitude)
@@ -51,8 +54,8 @@ extract_coordinates <- function(geo) {
       }
     }
   }
-  
-  return(NULL)  # Return NULL for invalid or empty data
+
+  return(NULL) # Return NULL for invalid or empty data
 }
 
 # Apply the function to extract all coordinates from the 'geometry' column
@@ -66,7 +69,7 @@ data$id <- seq_len(nrow(data))
 
 # Combine all valid coordinates into a single dataframe with IDs
 coordinates_data <- do.call(rbind, Map(function(coords, id) {
-  coords$id <- id  # Add ID to each set of coordinates
+  coords$id <- id # Add ID to each set of coordinates
   return(coords)
 }, valid_coordinates, data$id[!sapply(all_coordinates, is.null)]))
 
@@ -92,11 +95,11 @@ data <- read_csv("data/raw_data/raw_data.csv")
 
 # Select and clean the 'UPGRADED' column, then summarize by year
 upgraded_data <- data %>%
-  select(UPGRADED) %>%                     # Keep only the 'UPGRADED' column
-  filter(!is.na(UPGRADED) & UPGRADED > 1800) %>%  # Remove NA and invalid years
+  select(UPGRADED) %>% # Keep only the 'UPGRADED' column
[…]

### 2024-09-26 scripts/03-test_data.R

@@ -32,9 +32,9 @@ test_that("Test for missing values", {
 test_that("Test for valid data types", {
   # Check if INSTALLED, UPGRADED, OBJECTID are numeric
   data <- read_csv(here("data", "raw_data", "raw_data.csv"))
-  numeric_check <- all(sapply(data[c('INSTALLED', 'UPGRADED', 'OBJECTID')], is.numeric))
+  numeric_check <- all(sapply(data[c("INSTALLED", "UPGRADED", "OBJECTID")], is.numeric))
   expect_true(numeric_check, info = "INSTALLED, UPGRADED, OBJECTID should be numeric.")
-  
+
   # Check if LAST_EDIT_DATE has a valid date format (using ymd_hms for ISO 8601)
   date_check <- all(!is.na(ymd_hms(data$LAST_EDIT_DATE)))
   expect_true(date_check, info = "LAST_EDIT_DATE should be in valid date format.")
@@ -43,13 +43,13 @@ test_that("Test for valid data types", {
 test_that("Test for valid ranges of years", {
   data <- read_csv(here("data", "raw_data", "raw_data.csv"))
   current_year <- year(Sys.Date())
-  
+
   # Check INSTALLED range
   installed_valid <- all(data$INSTALLED >= 1900 & data$INSTALLED <= current_year, na.rm = TRUE)
   expect_true(installed_valid, info = "INSTALLED year should be between 1900 and the current year or NA.")
-  
+
   # Check UPGRADED range, allow NA values
-  upgraded_valid <- all(is.na(data$UPGRADED) | (data$UPGRADED >= 1900 & data$UPGRADED <= current_year | data$UPGRADED == 1)) #there is a invalid UPGRADED data 1
+  upgraded_valid <- all(is.na(data$UPGRADED) | (data$UPGRADED >= 1900 & data$UPGRADED <= current_year | data$UPGRADED == 1)) # there is a invalid UPGRADED data 1
   expect_true(upgraded_valid, info = "UPGRADED year should be between 1900 and the current year or NA.")
 })
 
@@ -58,7 +58,7 @@ test_that("Test for uniqueness of X_id and SEGMENT_ID", {
   # X_id uniqueness
   x_id_unique <- n_distinct(data$X_id) == nrow(data)
   expect_true(x_id_unique, info = "X_id should be unique.")
-  
+
   # SEGMENT_ID uniqueness
[…]

### 2024-09-26 scripts/04-charts.R

@@ -26,7 +26,7 @@ toronto_bbox <- c(-79.6303856025883, 43.5822069220627, -79.1180336515019, 43.855
 
 # Fetch map data from OpenStreetMap (OSM)
 toronto_map <- opq(bbox = toronto_bbox) %>%
-  add_osm_feature(key = 'highway', value = 'primary') %>%
+  add_osm_feature(key = "highway", value = "primary") %>%
   osmdata_sf()
 
 # Convert map data to Simple Features (sf) objects
@@ -34,18 +34,24 @@ streets <- toronto_map$osm_lines
 
 # Plot the map with bike lanes data
 ggplot() +
-  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) +  # OSM street map layer
-  geom_point(data = coordinates_data, aes(x = longitude, y = latitude), 
-             size = 0.1, color = "black") +  # Bike lanes points
-  labs(title = "Map of Bike Lanes Locations in Toronto", 
-       x = "Longitude", y = "Latitude") +
-  coord_sf(xlim = c(-79.6303856025883, -79.1180336515019), 
-           ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE) +
+  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) + # OSM street map layer
+  geom_point(
+    data = coordinates_data, aes(x = longitude, y = latitude),
+    size = 0.1, color = "black"
+  ) + # Bike lanes points
+  labs(
+    title = "Map of Bike Lanes Locations in Toronto",
+    x = "Longitude", y = "Latitude"
+  ) +
+  coord_sf(
+    xlim = c(-79.6303856025883, -79.1180336515019),
+    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
+  ) +
   theme_minimal() +
   theme(
-    panel.background = element_blank(),  # Remove panel background
-    panel.grid.major = element_blank(),  # Remove major grid lines
-    panel.grid.minor = element_blank(),  # Remove minor grid lines
+    panel.background = element_blank(), # Remove panel background
+    panel.grid.major = element_blank(), # Remove major grid lines
+    panel.grid.minor = element_blank(), # Remove minor grid lines
     axis.text.x = element_text(angle = 45, hjust = 1)
   )
 
@@ -61,28 +67,32 @@ upgraded_data <- read_csv("data/analysis_data/upgraded_data.csv")
 
 # Generate the bar chart for "Number of Bikeways Upgraded by Year"
[…]

### 2024-09-26 scripts/04-charts.R
chart map updated

@@ -27,6 +27,9 @@ toronto_bbox <- c(-79.6303856025883, 43.5822069220627, -79.1180336515019, 43.855
 # Fetch map data from OpenStreetMap (OSM)
 toronto_map <- opq(bbox = toronto_bbox) %>%
   add_osm_feature(key = "highway", value = "primary") %>%
+  add_osm_feature(key = 'highway', value = 'secondary') %>%
+  add_osm_feature(key = 'highway', value = 'tertiary') %>%
+  add_osm_feature(key = 'highway', value = 'residential') %>%
   osmdata_sf()
 
 # Convert map data to Simple Features (sf) objects

### 2024-09-26 paper/paper.qmd
paper reference updated
paper third draft

@@
 
-This project is structured based on Rohan's principles [@citeRohan].
+This project is motivated and guided by Rohan Alexander [@citeRohan] and his book [@citeTbook].
 Data used in this paper was downloaded, cleaned and analyzed with the programming language R [@citeR].
@@
 #| fig-cap: "Toronto Bikeways Installed by Year"
-#| fig-subcap: ["Number", "Trend"]
+#| fig-subcap: ["Number of bikeways installed", "Trend of bikeway installations"]
 #| echo: false
@@
 #| fig-cap: "Toronto Bikeways Upgraded by Year"
-#| fig-subcap: ["Number", "Trend"]
+#| fig-subcap: ["Number of bikeways upgraded", "Trend of bikeway upgrads"]
 #| echo: false
@@
 #| label: fig-map
-#| fig-cap: "Map of Bike Lanes Locations in Toronto"
+#| fig-cap: "Map of Toronto Bikeway Locations"
 #| echo: false
@@
   add_osm_feature(key = 'highway', value = 'primary') %>%
+  add_osm_feature(key = 'highway', value = 'secondary') %>%
+  add_osm_feature(key = 'highway', value = 'tertiary') %>%
+  add_osm_feature(key = 'highway', value = 'residential') %>%
   osmdata_sf()
@@
     axis.text.x = element_text(angle = 45, hjust = 1)
-  )
+  ) +
+  annotate("text", x = -79.12, y = 43.585, label = "Locations of bikeways in Toronto overlaid on a street map\
+           including primary, secondary, tertiary, and residential roads.", 
+           size = 5, hjust = 1, vjust = 0, color = "darkgrey")  # Caption at bottom right
 

### 2024-09-27 scripts/00-simulate_data.R
final draft

@@ -4,8 +4,6 @@
 # Date: 23 September 2024
 # Contact: ziheng.zhong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: NA
-# Any other information needed: NA
 
 
 

### 2024-09-27 scripts/04-charts.R

@@ -4,8 +4,7 @@
 # Date: 23 September 2024
 # Contact: ziheng.zhong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: NA
-# Any other information needed: NA
+
 
 
 #### Workspace setup ####

### 2024-09-27 paper/paper.qmd
paper final draft

@@
 
-The structure of this paper is organized as follows: @sec-data presents the data sources and methodologies, followed by an overview of the results in @sec-res.
+The structure of this paper is organized as follows: @sec-data presents the data sources, methodologies, and the visualization of the results.
 The findings are then discussed in @sec-dis, and additional detailed information is provided in the appendix at @sec-app.
@@
 
-# Results {#sec-res}
+## Charts
 
-## Data Trend
+### Data Trend
 
@@
 
-## Distributions
+### Distributions
 
@@
 
-## Maps
+### Maps
 
